### 2019-03-11 paper.tex
initial commit

\documentclass[11pt, titlepage]{article}
\usepackage{amssymb,amsmath,latexsym}
%\usepackage[spanish]{babel}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{hyperref}

% Page length commands go here in the preamble
\setlength{\oddsidemargin}{-0.25in} % Left margin of 1 in + 0 in = 1 in
\setlength{\textwidth}{7in}   % Right margin of 8.5 in - 1 in - 6.5 in = 1 in
\setlength{\topmargin}{-.75in}  % Top margin of 2 in -0.75 in = 1 in
\setlength{\textheight}{9.2in}  % Lower margin of 11 in - 9 in - 1 in = 1 in
\setlength{\parindent}{0in} 
\begin{document}
\title{\Huge{Versioned Datasets and Queries in RDF/SPARQL}}
\author{Ignacio Cuevas\\ Aidan Hogan}
\date{\today}
\maketitle
\newpage
%\abstract{} % Do we need one?
\section{Introduction}
The Semantic Web~\footnote{\url{https://www.w3.org/2001/sw/wiki/Main_Page}} is an initiative of the World Wide Web Consortium, which envisions a ``Web of Data;'' that is, a web of \textit{Linked Data} where information has an standarized structure, making it easier for systems and machines to comunicate. In order to make such standarization possible, it becomes necessary to incorporate meta-data that describes the content, meaning and relationship of data in all databases.

The Resource Description Framework~\cite{key:rdfprimer11} (RDF for short) is a standard model for data interchange on the Web. Its design is based on ``triples,'' which represent connections on a graph, where every node is a ``resource'' with ``predicates'' as the edges connecting them. Both resources and predicates are formally defined, and a collection of rules allow for inferences about the relationships in the data.
Alongside RDF, the ontological languages RDFS~\cite{key:rdfschema11} and OWL~\cite{key:owl2primer} provide basic relation and type definitions, and SPARQL~\cite{key:sparql11} serves as RDF's query engine (similar to SQL on relational databases).

Currently, a large number of existing datasets are based on RDF due to its extense popularity, Wikidata being one the most notorious. Wikidata~\cite{VrandecicK14} is a massive dataset, with over 42 million items~\footnote{\url{https://www.wikidata.org/wiki/Wikidata:Statistics}}, which is based on all the available data of Wikipedia and structured on RDF.

Wikidata is in constant change, due to it being an open dataset that allows being edited by users. Furthermore, data is being added often, which makes Wikidata very dynamic in some areas. Given those circumstances, it can be said that Wikidata has many ``versions'' over time, where different data is available at different time intervals, making it possible to analyze its time component.

Dataset versioning in RDF/SPARQL is a research topic yet to be fully explored. While there are a number of implementations, most of them rely on specialized indices or SPARQL extensions. As such, it would be relevant to study the possibility of developing a versioning system and methodology for queries using only base SPARQL.

Such system could be used as version control for the dataset; administrators would be able to detect erroneous or malicious editions on data that doesn't normally change. On top of that, it would be possible to analyze the evolution of data, possibly predicting future changes. The system could also be used on datasets other than Wikidata, where data history may be poorly (or not at all) preserved.

\section{Related Work}

\section{Preliminaries} % Maybe not
\section{(Method)} % Name of our method goes here
\section{Evaluation} % This two sections can
\section{Results}    % be merged for better flow
\section{Conclusion}
\newpage
\bibliographystyle{unsrt}
\bibliography{propuesta}
\end{document}

### 2019-03-12 paper.tex
added bibliography file and made some progress on the related work section

@@
 \section{Related Work}
+Gutierrez et al.~\cite{GutierrezHV07} describe a framework that allows time representation in RDF using annotations.
+An important distinction to make is that their work is focused on \textbf{labeling} over \textbf{versioning}; \textbf{labeling} means adding meta-data to query result validity, meanwhile \textbf{versioning} implies simultaneously keeping several versions of the same dataset.
+This work also formally defines labels and time intervals, but does not cover an implementation for either.
 
+Zimmerman et. al~\cite{DBLP:journals/ws/ZimmermannLPS12} extend RDF to allow generalized annotations which can, in particular, be used to add validity time intervals to each triple.
 \section{Preliminaries} % Maybe not
@@
 \bibliographystyle{unsrt}
-\bibliography{propuesta}
+\bibliography{paper}
 \end{document}

### 2019-03-12 paper.tex
Adding abstract, moving to ISWC template

@@
-\documentclass[11pt, titlepage]{article}
-\usepackage{amssymb,amsmath,latexsym}
-%\usepackage[spanish]{babel}
+\RequirePackage{amsmath}
+\documentclass{llncs}
 \usepackage[utf8]{inputenc}
-\usepackage{graphicx}
+\PassOptionsToPackage{table}{xcolor}
+\pagestyle{headings}  % This option shows page numbers, it have to be removed in the final document.
+\usepackage{booktabs}
+\usepackage{amssymb}
+\usepackage{tikz}
+\usepackage{scalerel}
+\usepackage{comment}
+\usepackage{framed}
+\usepackage{listings}
+\usepackage{pifont}% http://ctan.org/pkg/pifont
+\usepackage{balance}
+\usepackage{stmaryrd}
+\usepackage{adjustbox}
+\usepackage{multirow}
+\usepackage{enumitem}
+\usepackage{xspace}
+\usepackage{url}
+\usepackage{inconsolata}
+\usepackage{etoolbox}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Quickly switch from conference to extended version
+\providetoggle{conf}
+\settoggle{conf}{false}
+
+% text only to appear in extended version
+\newcommand{\ever}[1]{\iftoggle{conf}{}{#1}}
+
+% text only to appear in the conference version
+\newcommand{\cver}[1]{\iftoggle{conf}{#1}{}}
+
+% alternate text for both versions
+\newcommand{\cever}[2]{\iftoggle{conf}{#1}{#2}}
+
+% environment for long extended-only text
+\iftoggle{conf}%
+ {\excludecomment{exver}}%
+ {\newenvironment{exver}}%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\usepackage{floatrow}
+% Table float box with bottom caption, box width adjusted to content
+\newfloatcommand{capbtabbox}{table}[][\FBwidth]
+
+\newcommand{\ah}[1]{{\color{blue}\textsc{ah:} #1}}
+
+\makeatletter
+\renewcommand\paragraph{\@startsection{paragraph}{4}{\z@}%
+	{1ex \@plus1ex \@minus.2ex}%
+	{-1em}%
+	{\normalfont\normalsize\itshape}}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% SPARQL Listing style
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\usepackage{listings}
+
+\colorlet{punct}{red!60!black}
+\definecolor{background}{HTML}{EEEEEE}
+\definecolor{delim}{RGB}{120,20,40}
+\definecolor{keyw}{RGB}{0,0,192}
+\colorlet{numb}{magenta!60!black}
+
+\lstdefinelanguage{sparql}{
+	sensitive=false,
+	extendedchars=true,
+	literate={á}{{\'a}}1 {é}{{\'e}}1 {í}{{\'{\i}}}1 {ó}{{\'o}}1 {ú}{{\'u}}1
+	{Á}{{\'A}}1 {É}{{\'E}}1 {Í}{{\'I}}1 {Ó}{{\'O}}1 {Ú}{{\'U}}1
+	{ü}{{\"u}}1 {Ü}{{\"U}}1 {ñ}{{\~n}}1 {Ñ}{{\~N}}1 {¿}{{?``}}1 {¡}{{!``}}1
+	{<}{{{\color{delim}<}}}{1}
+	{>}{{{\color{delim}>}}}{1}
+	{?}{{{\color{delim}?}}}{1}
+	{*}{{{\color{delim}*}}}{1}
+	{+}{{{\color{delim}+}}}{1}
+	{/}{{{\color{delim}/}}}{1}
+	{,}{{{\color{punct}{,}}}}{1}
+	{;}{{{\color{punct}{;}}}}{1}
+	{.}{{{\color{punct}{.}}}}{1}
+	{:}{{{\color{punct}{:}}}}{1}
+	{\{}{{{\color{delim}{\{}}}}{1} {\}}{{{\color{delim}{\}}}}}{1},
+	morekeywords={ask,select,from,where,order,by,distinct,limit,offset,optional,union,filter,prefix,bound,desc,regex,str,group,not,exists,minus,service,certain,maybe}
+}
+
+\lstdefinestyle{sparqld}{
+	basicstyle=\scriptsize\ttfamily,
+	identifierstyle=\color{black},
+	keywordstyle=\color{keyw}\bfseries,
+	ndkeywordstyle=\color{greenCode}\bfseries,
+	stringstyle=\color{ocherCode}\ttfamily,
+	commentstyle=\color{darkgray}\ttfamily,
+	language={sparql},
+	tabsize=2,
+	showtabs=false,
+	showspaces=false,
+	showstringspaces=false,
+	extendedchars=true,
+	escapechar=`,
+	frame={single},
+	breaklines=true,
+	basewidth=0.5em,
+	moredelim=[is][\color{magenta}]{~}{~},
+	moredelim=**[is][\color{gray}]{£}{£},
+	moredelim=**[is][\color{blue!50!black}]{$}{$},
+	moredelim=[is][\color{orange!80!black}]{!}{!},
+	moredelim=**[is][\color{green!50!black}]{¬}{¬},
+	xleftmargin=2ex,
+	xrightmargin=1ex,
+	aboveskip=1.5ex,
+	belowskip=1.5ex
+}
+
+\makeatletter
+\newcommand{\sqbox}{%
+	\collectbox{%
+		\@tempdima=\dimexpr\width-\totalheight\relax
+		\ifdim\@tempdima<\z@
+		\fbox{\hbox{\hspace{-.5\@tempdima}\BOXCONTENT\hspace{-.5\@tempdima}}}%
+		\else
+		\ht\collectedbox=\dimexpr\ht\collectedbox+.5\@tempdima\relax
+		\dp\collectedbox=\dimexpr\dp\collectedbox+.5\@tempdima\relax
+		\fbox{\BOXCONTENT}%
+		\fi
+	}%
+}
+\makeatother
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% /SPARQL Listing style
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% TiKz for RDF graphs
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\usetikzlibrary{shapes,arrows,positioning,fit,backgrounds,matrix,chains,scopes,calc}
+
+\newcommand{\hsp}{\vphantom{Ag}}
+\tikzset{
+	std/.style={ 
+		draw,
+		circle,
+		anchor=center,
+		inner sep=0pt,
+		minimum size=5pt
+	},
+	lab/.style={ 
+		text centered,
+		fill=white, 
+		inner sep=0.7pt,
+		font=\tt\small\hsp},
+	iri/.style={
+		draw=black!50!white, 
+		rectangle,
+		rounded corners,
+		thick,
+		text centered,
+		top color=white, 
+		bottom color=black!15,
+		%opacity=0.7,
+		%text opacity=1,
+		font=\tt\small\hsp,
+		anchor=center},
+	lit/.style={
+		draw=black!50!white, 
+		rectangle,
+		thick,
+		text centered,
+		top color=white, 
+		bottom color=black!15, 
+		anchor=center,
+		%opacity=0.7,
+		%text opacity=1,
+		font=\tt\small\hsp},
+	arrout/.style={
+		->,
+		-latex,
+		%draw=black!50, 
+		%fill=black!50,
+		%thick,
+		font=\tt\small\hsp},
+	arrin/.style={
+		<-,
+		latex-,
+		%draw=black!50, 
+		%fill=black!50,
+		%thick,
+		font=\tt\small\hsp},
+	arrinout/.style={
+		<->,
+		latex-latex,
+		%draw=black!50, 
+		%fill=black!50,
+		%thick,
+		font=\tt\small\hsp},
+	arrstd/.style={
+		-,
+		%draw=black!50, 
+		%fill=black!50,
+		thick},
+	dashmed/.style={
+		-,
+		%draw=black!50, 
+		%fill=black!50,
+		thick,
+		dash pattern=on 3pt off 3pt,
+		font=\tt\small\hsp},
+	fade/.style={
+		opacity=0.4,
+		text opacity=0.4
+	},
+	fadet/.style={
+		opacity=1,
+		text opacity=0.4
+	},
+	every loop/.style={
+		<-,
+		latex-,
+		fill=black!50,
+		min distance=10mm,
+		in=0,
+		out=60,
+		looseness=10,
+		draw=black!50,
+		thick,
+		font=\tt\small\hsp
+	},
+	lean/.style={
+		dotted,
+		blue
+	},
+	leane/.style={
+		dotted,
+		blue
+	},
+	leanl/.style={
+		text=blue
+	}	
+}
+
+\newlength{\hgap}
+\newlength{\vgap}
+\setlength{\hgap}{2cm}
+\setlength{\vgap}{1.2cm}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% /TiKz for RDF graphs
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% PGFplots
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\usepackage{pgfplots}
+\usepgfplotslibrary{groupplots}
+\usepackage{pgfplotstable}
+\pgfplotsset{compat=newest}
+\usetikzlibrary{pgfplots.statistics}
+
+\makeatletter
+\pgfplotsset{
+	boxplot prepared from table/.code={
+		\def\tikz@plot@handler{\pgfplotsplothandlerboxplotprepared}%
+		\pgfplotsset{
+			/pgfplots/boxplot prepared from table/.cd,
+			#1,
+		}
+	},
+	/pgfplots/boxplot prepared from table/.cd,
+	table/.code={\pgfplotstablecopy{#1}\to\boxplot@datatable},
+	row/.initial=0,
+	make style readable from table/.style={
+		#1/.code={
+			\pgfplotstablegetelem{\pgfkeysvalueof{/pgfplots/boxplot prepared from table/row}}{##1}\of\boxplot@datatable
+			\pgfplotsset{boxplot/#1/.expand once={\pgfplotsretval}}
+		}
+	},
+	make style readable from table=lower whisker,
+	make style readable from table=upper whisker,
+	make style readable from table=lower quartile,
+	make style readable from table=upper quartile,
+	make style readable from table=median,
+}
+\makeatother
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% /PGFplots
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\usepackage{arydshln}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Hyperref
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \usepackage{hyperref}
+\definecolor{dark-blue}{rgb}{0.0,0.0,0.2}
+\definecolor{dark-green}{rgb}{0.0,0.2,0.0}
+\definecolor{dark-red}{rgb}{0.2,0.0,0.0}
+\hypersetup{
+	colorlinks, linkcolor={dark-red},
+	citecolor={dark-green}, urlcolor={dark-blue},
+	pdftitle={Canonicalisation of Monotone SPARQL Queries},    % title
+	pdfauthor={Jaime Salas, Aidan Hogan},     % author
+	pdfsubject={ISWC 2018},   % subject of the document
+	pdfkeywords={sparql;} {canonicalisation;} {containment;} {query;} {equivalence;} {caching;}, % list of keywords
+}
 
-% Page length commands go here in the preamble
-\setlength{\oddsidemargin}{-0.25in} % Left margin of 1 in + 0 in = 1 in
-\setlength{\textwidth}{7in}   % Right margin of 8.5 in - 1 in - 6.5 in = 1 in
-\setlength{\topmargin}{-.75in}  % Top margin of 2 in -0.75 in = 1 in
-\setlength{\textheight}{9.2in}  % Lower margin of 11 in - 9 in - 1 in = 1 in
-\setlength{\parindent}{0in} 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% /Hyperref
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%% MACROS
+
+\newcommand{\B}{\ensuremath{\mathbf{B}}\xspace}
+\newcommand{\I}{\ensuremath{\mathbf{I}}\xspace}
+\renewcommand{\L}{\ensuremath{\mathbf{L}}\xspace}
+\newcommand{\V}{\ensuremath{\mathbf{V}}\xspace}
+
+\newcommand{\cpx}[2]{\ensuremath{\textsc{#1-}\mathrm{#2}}\xspace}
+\newcommand{\npc}{\cpx{NP}{complete}}
+\newcommand{\nph}{\cpx{NP}{hard}}
+\newcommand{\psc}{\cpx{PSpace}{complete}}
+\newcommand{\psh}{\cpx{PSpace}{hard}}
+\newcommand{\gic}{\cpx{GI}{complete}}
+\newcommand{\gih}{\cpx{GI}{hard}}
+\newcommand{\dpc}{\cpx{DP}{complete}}
+\newcommand{\dph}{\cpx{DP}{hard}}
+\newcommand{\conpc}{\cpx{coNP}{complete}}
+\newcommand{\conph}{\cpx{coNP}{hard}}
+\newcommand{\ptp}{\ensuremath{\Pi_2^P}}
+\newcommand{\ptph}{\cpx{\ptp}{hard}}
+\newcommand{\ptpc}{\cpx{\ptp}{complete}}
+
+\newcommand{\tch}[1]{\textbf{#1}}
+\newcommand{\rid}[1]{\textsc{#1}}
+\newcommand{\ttl}[1]{\textsf{#1}}
+\newcommand{\tid}[1]{\textsc{#1}}
+%\newcommand{\dom}{\tid{dom}}
+%\newcommand{\rng}{\tid{rng}}
+%\newcommand{\sC}{\tid{sC}}
+%\newcommand{\sP}{\tid{sP}}
+
+\newcommand{\ssyn}[3]{[\ensuremath{#1\,\textsc{#2}\,#3}]}
+\newcommand{\sand}[2]{\ssyn{#1}{and}{#2}}
+\newcommand{\suni}[2]{\ssyn{#1}{union}{#2}}
+\newcommand{\sopt}[2]{\ssyn{#1}{opt}{#2}}
+\newcommand{\sfil}[2]{\ensuremath{\textsc{filter}_{#2}(#1)}}
+\newcommand{\ssel}[2]{\ensuremath{\textsc{select}_{#2}(#1)}}
+\newcommand{\sseld}[2]{\ensuremath{\textsc{select}^\Delta_{#2}(#1)}}
+\newcommand{\dom}[1]{\ensuremath{\mathrm{dom}(#1)}}
+\newcommand{\can}[1]{\ensuremath{\mathrm{can}(#1)}}
+\newcommand{\com}[2]{\ensuremath{#1 \sim #2}}
+
+%\newcommand{\hsc}[1]{{\footnotesize\MakeUppercase{#1}}}
+\newcommand{\hsc}[1]{#1}
+
+%\newcommand{\utp}[1]{\textsc{tp}{(#1)}}
+\newcommand{\ufo}[1]{\textsf{\hsc{\upshape #1}}}
+\newcommand{\uand}[1]{\ensuremath{\ufo{and}(#1)}}
+\newcommand{\uuni}[1]{\ensuremath{\ufo{union}(#1)}}
+\newcommand{\usel}[2]{\ensuremath{\ufo{select}_{#2}(#1)}}
+\newcommand{\useld}[2]{\ensuremath{\ufo{select}^\Delta_{#2}(#1)}}
+
+\newcommand{\uandn}{\ensuremath{\ufo{and}}}
+\newcommand{\uunin}{\ensuremath{\ufo{union}}}
+
+\newcommand{\bn}[1]{\texttt{\_:#1}}
+\newcommand{\iri}[1]{\texttt{:#1}}
+\newcommand{\var}[1]{\texttt{?#1}}
+
+\newcommand{\ican}[1]{\ensuremath{\textsc{iCan}(#1)}}
+\newcommand{\ecan}[1]{\ensuremath{\textsc{eCan}(#1)}}
+
+\newcommand{\ev}[2]{\ensuremath{#1(#2)}}
+
+\def\ojoin{\setbox0=\hbox{$\bowtie$}%
+	\rule[0.18ex]{.25em}{.5pt}\llap{\rule[.9ex]{.25em}{.5pt}}}
+\def\loj{\mathbin{\ojoin\mkern-5.8mu\bowtie}}
+
+\newcommand{\da}{\ensuremath{:\nolinebreak\mkern-1.2mu\nolinebreak=}}
+
+\newcommand{\qedr}{\begin{flushright}\qed\end{flushright}}
+
+\newcommand{\yt}{\ding{51}}%
+\newcommand{\nt}{\ding{55}}%
+
+\newcommand{\para}[1]{\smallskip\noindent\textbf{#1:}}
+
+\newcommand{\mq}{\textsc{mq}\xspace}
+\newcommand{\mqs}{\textsc{mq}s\xspace}
+\newcommand{\ucq}{\textsc{ucq}\xspace}
+\newcommand{\ucqs}{\textsc{ucq}s\xspace}
+\newcommand{\cq}{\textsc{cq}\xspace}
+\newcommand{\cqs}{\textsc{cq}s\xspace}
+\newcommand{\cuq}{\textsc{ucq}\xspace}
+\newcommand{\cuqs}{\textsc{ucq}s\xspace}
+
+%%%%%%%%
+
+\graphicspath{ {images/} }
+
 \begin{document}
-\title{\Huge{Versioned Datasets and Queries in RDF/SPARQL}}
-\author{Ignacio Cuevas\\ Aidan Hogan}
-\date{\today}
+\title{Versioned Queries in SPARQL}
+\author{Ignacio Cuevas \and Aidan Hogan}
+
+\institute{IMFD Chile \& Department of Computer Science, University of Chile}
+
 \maketitle
-\newpage
-%\abstract{} % Do we need one?
+
+\begin{abstract}
+Although many prominent Linked Datasets are in a state of constant evolution, historical data are rarely published in a standard way: while dumps are sometimes provided of old versions of the dataset, there is no widely agreed-upon way for clients to quickly see historical changes for the description of a given resource,  or the results of a given SPARQL query, etc.
+In this paper, we focus on the problem of indexing historical versions of an RDF graph in a SPARQL engine and rewriting queries according to two versioning primitives: viewing results from a particular version, or viewing changes between two consecutive versions.
+Aiming to maintain compatibility with off-the-shelf SPARQL engines, we identify and compare several models by which historical versions of an RDF graph can be stored as named graphs, and discuss how base queries can be rewritten in each model per the two aforementioned primitives.
+We then present a performance comparison of these models with respect to X weekly versions of Wikidata and Y user-defined queries with versioning, using Virtuoso and Jena as reference SPARQL implementations.
+Our results show ...
+\end{abstract}
+
 \section{Introduction}

### 2019-04-07 paper.tex
finished first draft of the related work section

@@
 
-Zimmerman et. al~\cite{DBLP:journals/ws/ZimmermannLPS12} extend RDF to allow generalized annotations which can, in particular, be used to add validity time intervals to each triple.
+Zimmerman et al.~\cite{DBLP:journals/ws/ZimmermannLPS12} extend RDF, allowing the use of generalized annotations, which can be used to represent time intervals and associated to each triple to represent their validity.
+Same as the previously mentioned work, the objectives seeked by Zimmerman et al.
+do not completely align with the present work, due to the fact that an RDF extension is used.
+It will, however, be considered to compare results.
+
+Grandi~\cite{Grandi10} proposes an extension for SPARQL, including a time component in queries.
+No implementation details are provided, however.
+The methodology used for building timed queries will be compared to the one this work defines.
+
+The prior three works share their focus on formal definitions over implementations.
+They also do not consider eficiency or scalability analysis and employ SPARQL extensions and/or specialized indices.
+In light of these facts, said works will be considered only to compare results for the most part.
+
+Lastly, Tappolet and Bernstein~\cite{TappoletB09} add a time component to RDF's syntax, proposing an eficient method to make SPARQL queries on it, as well.
+The dataset is annotated with time intervals corresponding to each triple's validity, and a specialized index compliments the query engine.
+Since this is an alternative solution to the explored challenge, it will be used to compare results, while also evaluating whether their specialized index is necessary and if similar results can be achieved using standar SPARQL.
 \section{Preliminaries} % Maybe not

### 2019-04-08 paper.tex
minor edits to abstract

--- 2019-03-12 paper.tex
@@
 \begin{document}
-\title{Versioned Queries in SPARQL}
+\title{Versioned Queries in Native SPARQL Stores}
 \author{Ignacio Cuevas \and Aidan Hogan}
@@
 \begin{abstract}
-Although many prominent Linked Datasets are in a state of constant evolution, historical data are rarely published in a standard way: while dumps are sometimes provided of old versions of the dataset, there is no widely agreed-upon way for clients to quickly see historical changes for the description of a given resource,  or the results of a given SPARQL query, etc.
-In this paper, we focus on the problem of indexing historical versions of an RDF graph in a SPARQL engine and rewriting queries according to two versioning primitives: viewing results from a particular version, or viewing changes between two consecutive versions.
-Aiming to maintain compatibility with off-the-shelf SPARQL engines, we identify and compare several models by which historical versions of an RDF graph can be stored as named graphs, and discuss how base queries can be rewritten in each model per the two aforementioned primitives.
-We then present a performance comparison of these models with respect to X weekly versions of Wikidata and Y user-defined queries with versioning, using Virtuoso and Jena as reference SPARQL implementations.
-Our results show ...
+Although many prominent Linked Datasets are in a state of constant evolution, historical data are rarely published in a standard way.
+Though a number of works have recently addressed the issue of querying and managing archives of versioned RDF data, most propose dedicated solutions, which incurs a certain deployment cost.
+We rather propose solutions for representing archives of versioned RDF data using the standard SPARQL language, allowing existing implementations to be used.
+For querying, we consider two versioning primitives: viewing results from a particular version, or viewing changes between two consecutive versions.
+We identify and compare several models by which historical versions of an RDF graph can be stored as named graphs, and discuss how base queries can be rewritten in each model per the two aforementioned primitives.
+We then present a performance comparison of these models and baseline (non-)versioned queries with respect to 23 weekly versions of Wikidata and 232 user-defined queries, using Virtuoso as a reference SPARQL implementation.
 \end{abstract}

### 2019-04-08 paper.tex
Hyperref

--- 2019-04-07 paper.tex
@@
 \usepackage{listings}
-\usepackage{pifont}% http://ctan.org/pkg/pifont
-\usepackage{balance}
-\usepackage{stmaryrd}
+%\usepackage{pifont}% http://ctan.org/pkg/pifont
+%\usepackage{balance}
+%\usepackage{stmaryrd}
 \usepackage{adjustbox}
@@
 	colorlinks, linkcolor={dark-red},
-	citecolor={dark-green}, urlcolor={dark-blue},
-	pdftitle={Canonicalisation of Monotone SPARQL Queries},    % title
-	pdfauthor={Jaime Salas, Aidan Hogan},     % author
-	pdfsubject={ISWC 2018},   % subject of the document
-	pdfkeywords={sparql;} {canonicalisation;} {containment;} {query;} {equivalence;} {caching;}, % list of keywords
+	citecolor={dark-green}, urlcolor={dark-blue}%,
+%	pdftitle={Canonicalisation of Monotone SPARQL Queries},    % title
+%	pdfauthor={Jaime Salas, Aidan Hogan},     % author
+%	pdfsubject={ISWC 2018},   % subject of the document
+%	pdfkeywords={sparql;} {canonicalisation;} {containment;} {query;} {equivalence;} {caching;}, % list of keywords
 }

### 2019-04-08 paper.tex
Small edits to intro and adding some more references

@@
 \begin{document}
-\title{Versioned Queries in Native SPARQL Stores}
-\author{Ignacio Cuevas \and Aidan Hogan}
+\title{Versioned Queries on Native SPARQL Storage}
+%\author{Ignacio Cuevas \and Aidan Hogan}
 
-\institute{IMFD Chile \& Department of Computer Science, University of Chile}
+%\institute{IMFD Chile \& Department of Computer Science, University of Chile}
 
@@
 Although many prominent Linked Datasets are in a state of constant evolution, historical data are rarely published in a standard way.
-Though a number of works have recently addressed the issue of querying and managing archives of versioned RDF data, most propose dedicated solutions, which incurs a certain deployment cost.
+Though a number of works have recently addressed the issue of querying and managing archives of versioned RDF data, most propose dedicated solutions, which incurs a high deployment cost.
 We rather propose solutions for representing archives of versioned RDF data using the standard SPARQL language, allowing existing implementations to be used.
@@
 \section{Introduction}
-The Semantic Web~\footnote{\url{https://www.w3.org/2001/sw/wiki/Main_Page}} is an initiative of the World Wide Web Consortium, which envisions a ``Web of Data;'' that is, a web of \textit{Linked Data} where information has an standarized structure, making it easier for systems and machines to comunicate.
-In order to make such standarization possible, it becomes necessary to incorporate meta-data that describes the content, meaning and relationship of data in all databases.
 
-The Resource Description Framework~\cite{key:rdfprimer11} (RDF for short) is a standard model for data interchange on the Web.
-Its design is based on ``triples,'' which represent connections on a graph, where every node is a ``resource'' with ``predicates'' as the edges connecting them.
-Both resources and predicates are formally defined, and a collection of rules allow for inferences about the relationships in the data.
-Alongside RDF, the ontological languages RDFS~\cite{key:rdfschema11} and OWL~\cite{key:owl2primer} provide basic relation and type definitions, and SPARQL~\cite{key:sparql11} serves as RDF's query engine (similar to SQL on relational databases).
+A key aspect of the Web is its dynamic nature, where documents are frequently updated, deleted and added.
+Likewise when we speak of the Semantic Web, it is important to consider that sources may be dynamic and RDF datasets are subject to change.
+The traditional Web and the Semantic Web thus share similar strengths and limitations regarding the dynamics of information.
+In terms of strengths, the flexible nature of both webs means that documents can be updated with little restriction and with little need for centralised coordination.
+On the other hand, in terms of limitations, neither web has built in support for preservation, nor for versioning; for example, on the traditional Web, the HTTP protocol does not permit requesting a past version of a webpage.
+For this reason, a number of specialised ``Web archives'' have emerged that attempt to capture and track different intermittent versions of documents on the Web, the most prominent of which is the Internet Archive~\cite{JaffeK09}.
+While similar techniques can be applied to Semantic Web documents containing RDF data -- simply archiving and thus preserving the syntax of the document itself -- the structured nature of such content means that an RDF archive can potentially do a lot more.
+
+A number of works have looked at archiving RDF content from the Web.
+One such example is the Dynamic Linked Data Observatory (DyLDO)~\cite{KaferAUOH13}, which has been tracking and archiving changes in a sample of around 100,000 RDF documents each week since 2013.
+Other works rather focus on aspects relating to detecting change~\cite{TummarelloMBE07,ZeginisTC11,PapavasileiouFFKC13,KaferAUOH13,DividinoKG14,RoussakisCSFS15,NishiokaS18}, notifying of changes~\cite{TummarelloMBE07,PopitschH11,PassantM10,TrampFEA10}, versioning data~\cite{VolkelG06,GraubeHU14,KhuranaD16}, and so forth.
+
+% RDF archives~\cite{FernandezUPK19}
 
@@
 The prior three works share their focus on formal definitions over implementations.
-They also do not consider eficiency or scalability analysis and employ SPARQL extensions and/or specialized indices.
+They also do not consider efficiency or scalability analysis and employ SPARQL extensions and/or specialized indices.
 In light of these facts, said works will be considered only to compare results for the most part.

### 2020-09-04 paper.tex
changing title, adding authors

@@
 	colorlinks, linkcolor={dark-red},
-	citecolor={dark-green}, urlcolor={dark-blue}%,
-%	pdftitle={Canonicalisation of Monotone SPARQL Queries},    % title
-%	pdfauthor={Jaime Salas, Aidan Hogan},     % author
-%	pdfsubject={ISWC 2018},   % subject of the document
-%	pdfkeywords={sparql;} {canonicalisation;} {containment;} {query;} {equivalence;} {caching;}, % list of keywords
+	citecolor={dark-green}, urlcolor={dark-blue},
+	pdftitle={Versioned Queries over RDF Archives using Native SPARQL Storage},    % title
+	pdfauthor={Ignacio Cuevas, Aidan Hogan},     % author
+	pdfsubject={MEPDaW 2020: Managing the Evolution and Preservation of the Data Web},   % subject of the document
+	pdfkeywords={sparql;} {versioning;} {rdf archives;} {dynamics}, % list of keywords
 }
@@
 \begin{document}
-\title{Versioned Queries on Native SPARQL Storage}
-%\author{Ignacio Cuevas \and Aidan Hogan}
+\title{Versioned Queries over RDF Archives using Native SPARQL Storage}
 
-%\institute{IMFD Chile \& Department of Computer Science, University of Chile}
+\author{Ignacio Cuevas \and Aidan Hogan}
+\institute{Department of Computer Science, University of Chile \& IMFD Chile}
 

### 2020-09-07 paper.tex
update intro and related work

--- 2019-04-08 paper.tex
@@
 	colorlinks, linkcolor={dark-red},
-	citecolor={dark-green}, urlcolor={dark-blue}%,
-%	pdftitle={Canonicalisation of Monotone SPARQL Queries},    % title
-%	pdfauthor={Jaime Salas, Aidan Hogan},     % author
-%	pdfsubject={ISWC 2018},   % subject of the document
-%	pdfkeywords={sparql;} {canonicalisation;} {containment;} {query;} {equivalence;} {caching;}, % list of keywords
+	citecolor={dark-green}, urlcolor={dark-blue},
+	pdftitle={Versioned Queries over RDF Archives: All You Need is SPARQL?},    % title
+	pdfauthor={Ignacio Cuevas, Aidan Hogan},     % author
+	pdfsubject={MEPDaW 2020: Managing the Evolution and Preservation of the Data Web},   % subject of the document
+	pdfkeywords={sparql;} {versioning;} {rdf archives;} {dynamics}, % list of keywords
 }
@@
 \begin{document}
-\title{Versioned Queries in Native SPARQL Stores}
+\title{Versioned Queries over RDF Archives:\\All You Need is SPARQL?}
+
 \author{Ignacio Cuevas \and Aidan Hogan}
-
-\institute{IMFD Chile \& Department of Computer Science, University of Chile}
+\institute{Department of Computer Science, University of Chile \& IMFD Chile}
 
@@
 Although many prominent Linked Datasets are in a state of constant evolution, historical data are rarely published in a standard way.
-Though a number of works have recently addressed the issue of querying and managing archives of versioned RDF data, most propose dedicated solutions, which incurs a certain deployment cost.
-We rather propose solutions for representing archives of versioned RDF data using the standard SPARQL language, allowing existing implementations to be used.
-For querying, we consider two versioning primitives: viewing results from a particular version, or viewing changes between two consecutive versions.
+Though a number of works have recently addressed the issue of querying and managing archives of versioned RDF data, most propose dedicated solutions, which incurs a high deployment cost.
+We rather explore solutions for representing archives of versioned RDF data using the SPARQL standard without modification, allowing existing implementations to be used.
+For querying, we consider two versioning primitives: viewing results from a particular version, or viewing results that change between two consecutive versions.
 We identify and compare several models by which historical versions of an RDF graph can be stored as named graphs, and discuss how base queries can be rewritten in each model per the two aforementioned primitives.
@@
 \section{Introduction}
-The Semantic Web~\footnote{\url{https://www.w3.org/2001/sw/wiki/Main_Page}} is an initiative of the World Wide Web Consortium, which envisions a ``Web of Data;'' that is, a web of \textit{Linked Data} where information has an standarized structure, making it easier for systems and machines to comunicate.
-In order to make such standarization possible, it becomes necessary to incorporate meta-data that describes the content, meaning and relationship of data in all databases.
 
-The Resource Description Framework~\cite{key:rdfprimer11} (RDF for short) is a standard model for data interchange on the Web.
-Its design is based on ``triples,'' which represent connections on a graph, where every node is a ``resource'' with ``predicates'' as the edges connecting them.
-Both resources and predicates are formally defined, and a collection of rules allow for inferences about the relationships in the data.
-Alongside RDF, the ontological languages RDFS~\cite{key:rdfschema11} and OWL~\cite{key:owl2primer} provide basic relation and type definitions, and SPARQL~\cite{key:sparql11} serves as RDF's query engine (similar to SQL on relational databases).
+A key aspect of the Web is its dynamic nature, where documents are frequently updated, deleted and added.
+Likewise when we speak of the Semantic Web, it is important to consider that sources may be dynamic and RDF datasets are subject to change~\cite{KaferAUOH13}.
+It is in this context that various works have looked at versioning in the context of RDF/SPARQL~\cite{VolkelG06,TappoletB09,Grandi10,GraubeHU14,KhuranaD16}, with recent works proposing \textit{RDF archives}~\cite{FernandezPU15,Cerdeira-PenaFF16,BahriLA18,FernandezUPK19,TaelmanSHMV19} that manage RDF graphs and their historical changes, allowing for querying across different versions of the graph.
+Within these works, a variety of specialised indexing techniques~\cite{Cerdeira-PenaFF16,BahriLA18,TaelmanSHMV19}, query languages~\cite{TappoletB09} and benchmarks~\cite{KotsevMPEFK16,FernandezUPK19} have been proposed, developed and evaluated.
+While these represent important advances, many such proposals are by their nature incompatible with the existing SPARQL infrastructure.
+Furthermore, benchmarks have mostly focused on synthetic data.
+Hence there is still a considerable gap between these research works and putting RDF archives into practice.
 
-Currently, a large number of existing datasets are based on RDF due to its extense popularity, Wikidata being one the most notorious.
+%The traditional Web and the Semantic Web thus share similar strengths and limitations regarding the dynamics of information.
+In terms of strengths, the flexible nature of both webs means that documents can be updated with little restriction and with little need for centralised coordination.
+On the other hand, in terms of limitations, neither web has built in support for preservation, nor for versioning; for example, on the traditional Web, the HTTP protocol does not permit requesting a past version of a webpage.
+For this reason, a number of specialised ``Web archives'' have emerged that attempt to capture and track different intermittent versions of documents on the Web, the most prominent of which is the Internet Archive~\cite{JaffeK09}.
+While similar techniques can be applied to Semantic Web documents containing RDF data -- simply archiving and thus preserving the syntax of the document itself -- the structured nature of such content means that an RDF archive can potentially do a lot more.
+
+%A number of works have looked at tracking the dynamics of RDF data on the Web.
+One such example is the Dynamic Linked Data Observatory (DyLDO)~\cite{KaferAUOH13}, which has been tracking and archiving weekly changes in a diverse sample of RDF documents since 2013.
+
+In order to bridge this gap, we highlight that -- in theory at least -- there is no need for specialised indexes, query languages, etc., but rather that the types of versioned queries proposed in the literature~\cite{FernandezPU15} can be supported using off-the-shelf SPARQL engines that already enjoy years of development, optimisation, etc., as well as broad deployment on the Web.
+For example, SPARQL named graphs can be used to track different versions of individual graphs.
+However, as Fernandez at al.~\cite{FernandezUPK19} note, the approach of using pure SPARQL would ``\textit{typically render rather inefficient SPARQL queries}''.
+This claim raises the research question: how inefficient will pure SPARQL be in this setting?
+Though Fernandez at al.~\cite{FernandezUPK19} do discuss how pure SPARQL could be used, to the best of our knowledge, this question has not been empirically explored.
+If a pure SPARQL solution could be found with relatively little overhead versus base queries (over a single current version) then the benefits would be significant in terms of being able to leverage existing SPARQL infrastructure -- i.e., languages, protocols, implementations -- for managing and querying RDF archives.
+
+In this paper, we present preliminary empirical results addressing this research question.
+Specifically we take the Virtuoso SPARQL engine~\cite{Erling12} and  evaluate five different representations of RDF archives for XX different weekly versions of truthy Wikidata dumps~\cite{VrandecicK14}.
+We then implement query rewriting mechanisms for each representation in order to support single-version queries and sequential single-delta queries that retrieve all solutions over an indicated version and all solutions that differ between two sequential versions, respectively.
+We then perform experiments to compare the sizes of indexes, the costs of indexing new versions, and the costs of query evaluation for base queries over the current version versus rewritten versioned queries.
+
+%Other works rather focus on aspects relating to detecting change~\cite{TummarelloMBE07,ZeginisTC11,PapavasileiouFFKC13,KaferAUOH13,DividinoKG14,RoussakisCSFS15,NishiokaS18}, notifying changes~\cite{TummarelloMBE07,PopitschH11,PassantM10,TrampFEA10}, versioning data~\cite{VolkelG06,GraubeHU14,KhuranaD16}, and so forth.
+A more recent development is that of RDF archives~\cite{FernandezPU15,Cerdeira-PenaFF16,FernandezUPK19}
+
+%Currently, a large number of existing datasets are based on RDF due to its extense popularity, Wikidata being one the most notorious.
 Wikidata~\cite{VrandecicK14} is a massive dataset, with over 42 million items~\footnote{\url{https://www.wikidata.org/wiki/Wikidata:Statistics}}, which is based on all the available data of Wikipedia and structured on RDF.
-
-Wikidata is in constant change, due to it being an open dataset that allows being edited by users.
+%
+%
+%Wikidata is in constant change, due to it being an open dataset that allows being edited by users.
 Furthermore, data is being added often, which makes Wikidata very dynamic in some areas.
 Given those circumstances, it can be said that Wikidata has many ``versions'' over time, where different data is available at different time intervals, making it possible to analyze its time component.
-
-Dataset versioning in RDF/SPARQL is a research topic yet to be fully explored.
+%
+%
+%Dataset versioning in RDF/SPARQL is a research topic yet to be fully explored.
 While there are a number of implementations, most of them rely on specialized indices or SPARQL extensions.
 As such, it would be relevant to study the possibility of developing a versioning system and methodology for queries using only base SPARQL.
-
-Such system could be used as version control for the dataset; administrators would be able to detect erroneous or malicious editions on data that doesn't normally change.
+%
+%
+%Such system could be used as version control for the dataset; administrators would be able to detect erroneous or malicious editions on data that doesn't normally change.
 On top of that, it would be possible to analyze the evolution of data, possibly predicting future changes.
@@
 \section{Related Work}
+
+Various temporal extensions for RDF/SPARQL have been proposed in literature.
 Gutierrez et al.~\cite{GutierrezHV07} describe a framework that allows time representation in RDF using annotations.
-An important distinction to make is that their work is focused on \textbf{labeling} over \textbf{versioning}; \textbf{labeling} means adding meta-data to query result validity, meanwhile \textbf{versioning} implies simultaneously keeping several versions of the same dataset.
+Pugliese et al.~\cite{PuglieseUS08} more generally explore time annotations in RDF, proposing semantics, queries and specialised indexes.
+Zimmerman et al.~\cite{ZimmermannLPS12} provide a general framework for annotating RDF that generalises temporal annotations; they further propose a query language called AnQL for querying annotated RDF.
+Proposed temporal extensions for SPARQL include $\tau$-SPARQL~\cite{TappoletB09}, T-SPARQL~\cite{Grandi10}, SPARQL-ST~\cite{PerryJS11}, SPARQL\textsuperscript{T}~\cite{ZanioloGACG18}, etc.
+Related to temporality, a number of systems propose versioning supported for RDF, including SemVersion~\cite{VolkelG06}, x-RDF-3x~\cite{NeumannW10}, R43ples~\cite{GraubeHU14}, Dydra~\cite{AndersonB16} and Ostrich~\cite{TaelmanSV18}.
+
+More recently the concept of RDF archives have been gaining more and more attention.
+Fernandez et al.~\cite{FernandezPU15} provide a general survey and introduction to the theme, discussing the types of queries that one might wish to answer in the context of an RDF archive.
+Works by Cerdeira{-}Pena et al.~\cite{Cerdeira-PenaFF16}, Zaniolo et al.~\cite{ZanioloGACG18} and Taelman et al.~\cite{TaelmanSHMV19} propose ways to compress and index RDF archives, while Khurana and Deshpande~\cite{KhuranaD16} propose indexes more generally for historical graph data.
+Bahri et al.~\cite{BahriLA18} explore the benefits of using Apache Spark to manage RDF archives in a distributed setting.
+In order to evaluate these emerging techniques, a number of benchmarks have been proposed, including the BEnchmark of RDF
+ARchives (BEAR)~\cite{FernandezUPK19}, and the Semantic Publishing Benchmark (SPB)~\cite{Papakonstantinou18}.
+
+In summary, the past years have seen a wide range of developments in terms of representations, languages, vocabularies, indexes, algorithms, etc., for managing and querying over temporal and/or versioned RDF data.
+However, to the best of our knowledge, most of these approaches propose specialised languages, implementations, etc., for these purposes.
+Such languages have the benefit of allowing to more concisely and intuitively express versioned queries, while such implementations can develop optimised techniques that leverage particular characteristics of versioned data.
+The downside of these specialised approaches is that they are often diverge from practice, requiring the use of research prototypes rather than established SPARQL implementations, non-standard languages, etc.
+A number of authors have acknowledged that one can manage and query RDF archives using vanilla SPARQL, though it may lead to prolix or inefficient queries~\cite{TappoletB09,FernandezUPK19}; however, we are not aware of any empirical results along these lines.
+Our work presents some preliminary results that aim to fill this gap and to better understand the costs and limitations of using vanilla SPARQL to host RDF archives.
+
+%An important distinction to make is that their work is focused on \textbf{labeling} over \textbf{versioning}; \textbf{labeling} means adding meta-data to query result validity, meanwhile \textbf{versioning} implies simultaneously keeping several versions of the same dataset.
 This work also formally defines labels and time intervals, but does not cover an implementation for either.
 
-Zimmerman et al.~\cite{DBLP:journals/ws/ZimmermannLPS12} extend RDF, allowing the use of generalized annotations, which can be used to represent time intervals and associated to each triple to represent their validity.
+%Zimmerman et al.~\cite{DBLP:journals/ws/ZimmermannLPS12} extend RDF, allowing the use of generalized annotations, which can be used to represent time intervals and associated to each triple to represent their validity.
 Same as the previously mentioned work, the objectives seeked by Zimmerman et al.
@@
 
-Grandi~\cite{Grandi10} proposes an extension for SPARQL, including a time component in queries.
+%Grandi~\cite{Grandi10} proposes an extension for SPARQL, including a time component in queries.
 No implementation details are provided, however.
@@
 
-The prior three works share their focus on formal definitions over implementations.
-They also do not consider eficiency or scalability analysis and employ SPARQL extensions and/or specialized indices.
+%The prior three works share their focus on formal definitions over implementations.
+They also do not consider efficiency or scalability analysis and employ SPARQL extensions and/or specialized indices.
 In light of these facts, said works will be considered only to compare results for the most part.
 
-Lastly, Tappolet and Bernstein~\cite{TappoletB09} add a time component to RDF's syntax, proposing an eficient method to make SPARQL queries on it, as well.
+%Lastly, Tappolet and Bernstein~\cite{TappoletB09} add a time component to RDF's syntax, proposing an eficient method to make SPARQL queries on it, as well.
 The dataset is annotated with time intervals corresponding to each triple's validity, and a specialized index compliments the query engine.
 Since this is an alternative solution to the explored challenge, it will be used to compare results, while also evaluating whether their specialized index is necessary and if similar results can be achieved using standar SPARQL.
+
+% \cite{ArndtNRMM19} GIT for RDF
+
+% \cite{Papakonstantinou18} Benchmark results
+
+% \cite{PelgrinGH20}
+
 \section{Preliminaries} % Maybe not

### 2020-09-08 paper.tex
Adding formal definitions. :/

@@
 \usepackage{listings}
+\usepackage{mathtools}
 %\usepackage{pifont}% http://ctan.org/pkg/pifont
@@
 \newcommand{\sopt}[2]{\ssyn{#1}{opt}{#2}}
+\newcommand{\sminus}[2]{\ssyn{#1}{minus}{#2}}
 \newcommand{\sfil}[2]{\ensuremath{\textsc{filter}_{#2}(#1)}}
-\newcommand{\ssel}[2]{\ensuremath{\textsc{select}_{#2}(#1)}}
-\newcommand{\sseld}[2]{\ensuremath{\textsc{select}^\Delta_{#2}(#1)}}
+\newcommand{\ssel}[2]{\ensuremath{\textsc{select}_{#2}#1}}
+\newcommand{\sseld}[2]{\ensuremath{\textsc{select}^\Delta_{#2}#1}}
+\newcommand{\sfrom}[3]{\ensuremath{\textsc{from}_{#2,#3}#1}}
 \newcommand{\dom}[1]{\ensuremath{\mathrm{dom}(#1)}}
+\newcommand{\vars}[1]{\ensuremath{\mathrm{vars}(#1)}}
 \newcommand{\can}[1]{\ensuremath{\mathrm{can}(#1)}}
@@
 
-\section{Preliminaries} % Maybe not
-\section{(Method)} % Name of our method goes here
-\section{Evaluation} % This two sections can
-\section{Results}    % be merged for better flow
+\section{Preliminaries}
+
+We first establish some preliminaries and notation.
+RDF triples are composed of three sets of terms: IRIs (\I), literals (\L) and blank nodes (\B).
+In this preliminary work, we do not consider blank nodes, whose special semantics complicates the detection of changes in RDF graphs~\cite{ZeginisTC11}.
+Leaving aside blank nodes, an RDF triple $(s,p,o) \in \I \times \I \times (\I \cup \L)$ consists of a subject $s$, predicate $p$ and object $o$.
+An RDF graph $G$ is then a set of triples.
+An RDF archive is a tuple of RDF graphs $\mathcal{G} = (G_1,\ldots,G_n)$, interpreted as $n$ sequential versions of a given RDF graph.
+We denote by $\Delta_{i}^j \coloneqq G_i \setminus G_j$ the triples in version $i$ not in version $j$, and by $\mathcal{G}(s,p,o) \coloneqq \{ i \mid (s,p,o) \in G_i\text{ for } 1 \leq i \leq n \}$ the versions containing $(s,p,o)$.\footnote{The definition $\mathcal{G} : \I \times \I \times (\I \cup \L) \rightarrow 2^{\{1,\ldots,n\}}$~\cite{FernandezUPK19} is analogous to $\mathcal{G} = (G_1,\ldots,G_n)$.}
+
+A triple pattern $(s,p,o) \in (\I \cup \V) \times (\I \cup \V) \times (\I \cup \L \cup \V)$ is an RDF triple that permits variables from a set $\V$ to appear in any position.
+A solution is a partial mapping $\mu : \V \rightarrow \I \cup \L$.
+We denote by $\dom{\mu}$ the \textit{domain} of $\mu$, i.e., the set of variables for which $\mu$ is defined.
+We say that two solutions $\mu_1, \mu_2$ are compatible, denoted $\mu_1 \sim \mu_2$, if and only if $\mu_1(v) = \mu_2(v)$ for all $v \in \dom{\mu_1} \cap \dom{\mu_2}$.
+
+A basic graph pattern $B$ is a set of triple patterns.
+Let $\vars{B}$ denote the set of variables appearing in $B$ and let $\mu(B)$ denote the image of $B$ under $\mu$, i.e., the result of replacing all variables $v \in \dom{\mu} \cap \vars{B}$ with $\mu(v)$ in $B$.
+The evaluation of a basic graph pattern $B$ on an RDF graph $G$ is given as $B(G) \coloneqq \{ \mu \mid \mu(B) \subseteq G\text{ and }\dom{\mu} = \vars{B} \}$.
+
+SPARQL queries are based on basic graph patterns and a number of relational operators.
+Similar to Pérez et al.~\cite{PerezAG09}, we define an abstract syntax for a subset of SPARQL of pertinence to this paper as follows.
+A basic graph pattern $B$ is a graph pattern.
+Furthermore, if $P$ and $Q$ are graph patterns, and $V \subset \V$ is a set of variables, then $\sand{P}{Q}$, $\suni{P}{Q}$, $\sminus{P}{Q}$ and $\ssel{P}{V}$ are graph patterns.
+We define the semantics of these graph patterns in Table~\ref{tab:sparql}.
+
+\begin{table}[t]
+{\scriptsize
+\begin{align*}
+\sand{P}{Q}(G) \da &\, P(G) \Join Q(G) & M_1 \Join M_2 \da &\, \{ \mu_1 \cup \mu_2 \mid \mu_1 \in M_1,\, \mu_2 \in M_2,\, \com{\mu_1}{\mu_2} \} \\
+\suni{P}{Q}(G) \da &\, P(G) \cup Q(G) & M_1 \cup M_2 \da &\, \{ \mu \mid \mu \in M_1\text{ or }\mu \in M_2 \} \\
+\sminus{P}{Q}(G) \da &\, P(G) \setminus Q(G) &  M_1 \setminus M_2 \da &\, \{ \mu_1 \in M_1 \mid \nexists \mu_2 \in M_2 : \com{\mu_1}{\mu_2} \} \\
+%M_1 \loj M_2 \da\, & (M_1 \bowtie M_2) \cup (M_1 \setminus M_2) \\
+%\sigma_R(M) \da\,& \{ \mu \in M \mid \mu \models R \}\\
+\ssel{P}{V}(G) \da &\, \pi_V(P(G)) & \pi_V(M) \da\,& \{ \mu' \mid \exists \mu \in M : \mu' \sim \mu, \dom{\mu'} = V \cap \dom{\mu} \}
+\end{align*}}
+\caption{Semantics for SPARQL query operators where $M$, $M_1$ and $M_2$ denote sets of solutions, $V$ denotes a set of variables\label{tab:sparql}}
+\end{table}
+
+In this paper, we rely not only on RDF graphs, but also on SPARQL datasets to represent RDF archives.
+A SPARQL dataset $D \coloneqq \{ G, (n_1,G_1), \ldots (n_m,G_m) \}$ consists of a default (RDF) graph $G$ and a set of named graphs, each being a pair $(n_i,G_i)$ where $n_i \in \I$, $G_i$ is an RDF graph, and $n_i \neq n_j$ for $1 \leq i \leq m$, $1 \leq j \leq m$, $i \neq j$.
+For simplicity, we will consider a SPARQL dataset $D$ to be a set of quads of the form $(G \times \{*\}) \cup (G_1 \times \{n_1\}) \cup \ldots \cup (G_m \times \{n_m\})$, where $* \notin \I \cup \L \cup \V$ is a special symbol denoting the default graph.\footnote{We thus assume a \textit{quad store}, and disallow empty named graphs.} Different named graphs can be queried using a \texttt{GRAPH} operator, creating quad patterns.
+A quad pattern $(s,p,o,g) \in (\I \cup \V) \times (\I \cup \V) \times (\I \cup \L \cup \V) \times (\I \cup \{ *\})$ extends a triple pattern with a fourth element that may be an IRI or *.
+A basic named graph pattern $C$ is then a set of quad patterns whose evaluation on an SPARQL dataset $D$ is given as $C(D) \coloneqq \{ \mu \mid \mu(C) \subseteq D\text{ and }\dom{\mu} = \vars{C} \}$.
+Following the SPARQL standard~\cite{sparql11}, we translate a triple pattern $(s,p,o)$ to a quad pattern $(s,p,o,*)$ such that it is evaluated only on the default graph.\footnote{In concrete SPARQL syntax, quad patterns are expressed with the $\texttt{GRAPH}$ keyword.} The semantics of the operators defined in Table~\ref{tab:sparql} then remain unchanged simply allowing $P$ and $Q$ to now represent basic named graph patterns, defined to be graph patterns.
+
+SPARQL provides two operators to initialise a SPARQL dataset: \texttt{FROM} and \texttt{FROM NAMED}.
+We combine both for brevity into one operator.
+A graph pattern $P$ is considered to be a query.
+Likewise if $P$ is a graph pattern, and $M$ and $N$ are sets of IRIs, then $\sfrom{P}{M}{N}$ is also a query.
+The evaluation of $\sfrom{P}{M}{N}$ on a dataset $D$ is defined as $\sfrom{P}{M}{N}(D) \coloneqq P(D_{M,N})$, where the query dataset $D_{M,N} \coloneqq (\cup_{m \in M} D(m) \times \{ *\}) \cup  (\cup_{n \in N} D(n) \times n)$ is formed from a default graph containing the union\footnote{Since we do not allow blank nodes, a union or RDF merge is equivalent.} of graphs with a name $m \in M$, and all the named graphs in $D$ with a graph named $n \in N$.
+We will be able to use these operators to define query datasets that capture specific versions in an RDF archive.
+
+\section{Versioned Data}
+
+Our general approach is to represent an RDF archive as a SPARQL dataset $D$.
+
+\section{Versioned Queries}
+
+\section{Experiments}
+
 \section{Conclusion}

### 2020-09-08 paper.tex
adding description of data representations

@@
 Proposed temporal extensions for SPARQL include $\tau$-SPARQL~\cite{TappoletB09}, T-SPARQL~\cite{Grandi10}, SPARQL-ST~\cite{PerryJS11}, SPARQL\textsuperscript{T}~\cite{ZanioloGACG18}, etc.
-Related to temporality, a number of systems propose versioning supported for RDF, including SemVersion~\cite{VolkelG06}, x-RDF-3x~\cite{NeumannW10}, R43ples~\cite{GraubeHU14}, Dydra~\cite{AndersonB16} and Ostrich~\cite{TaelmanSV18}.
+Related to temporality, a number of systems support versioning for RDF, including SemVersion~\cite{VolkelG06}, POI~\cite{TzitzikasTA08}, x-RDF-3x~\cite{NeumannW10}, R43ples~\cite{GraubeHU14}, Dydra~\cite{AndersonB16} and Ostrich~\cite{TaelmanSV18}.
 
@@
 An RDF archive is a tuple of RDF graphs $\mathcal{G} = (G_1,\ldots,G_n)$, interpreted as $n$ sequential versions of a given RDF graph.
-We denote by $\Delta_{i}^j \coloneqq G_i \setminus G_j$ the triples in version $i$ not in version $j$, and by $\mathcal{G}(s,p,o) \coloneqq \{ i \mid (s,p,o) \in G_i\text{ for } 1 \leq i \leq n \}$ the versions containing $(s,p,o)$.\footnote{The definition $\mathcal{G} : \I \times \I \times (\I \cup \L) \rightarrow 2^{\{1,\ldots,n\}}$~\cite{FernandezUPK19} is analogous to $\mathcal{G} = (G_1,\ldots,G_n)$.}
 
@@
 
-Our general approach is to represent an RDF archive as a SPARQL dataset $D$.
+Our general approach is to represent an RDF archive as a SPARQL dataset $D$ but there are multiple representations by which this can be achieved, each with its own strength and weaknesses.
+We propose six different representations falling into three different categories as discussed by various authors (e.g.,~\cite{TzitzikasTA08,FernandezPU15}): \textit{Independent Copies} (\textit{IC}), \textit{Change-Based} (\textit{CB}) and \textit{Timestamp-Based} (\textit{TB}).
+
+\paragraph{Independent Copies (IC):} Perhaps the most intuitive representation is to store each constituent graph of an RDF Archive $\mathcal{G} = (G_1,\ldots,G_n)$ as a SPARQL dataset $D = \{ (x_1,G_1), \ldots, (x_n,G_n)\}$, where $x_1,\ldots,x_n$ are IRIs that deterministically identify the version.
+This will result in relatively simple (and thus likely efficient) rewritten queries, but can be expected to occupy a lot of space, particularly in cases where a small proportion of triples change from version to version.
+We thus consider the $IC$ representation to be a baseline for comparison.
+
+\paragraph{Change-Based (CB):} The core idea of $CB$ representations is to store only triples that change from a given reference version.
+Along these lines, in the following we denote by $\Delta_{i}^j \coloneqq G_i \setminus G_j$ the triples in version $i$ not in version $j$.
+We then consider four $CB$ representations based on the following transformations of an RDF archive $\mathcal{G} = (G_1,\ldots,G_n)$:
+%
+{\footnotesize
+\begin{align*}
+\mathcal{G}^n_1 \da\,& (G_1,\Delta_1^2,\Delta_2^1,\ldots,\Delta_1^n,\Delta_n^1) & G_i & = (G_1 \cup \Delta_i^1) \setminus \Delta_1^i & \text{for }1 < i \leq n\\
+\mathcal{G}^n_{n-1} \da\,& (G_1,\Delta_1^2,\Delta_2^1,\ldots,\Delta_{n-1}^n,\Delta_n^{n-1}) & G_i & = (G_{i-1} \cup \Delta_i^{i-1}) \setminus \Delta_{i-1}^i & \text{for }1 < i \leq n \\
+\mathcal{G}^1_n \da\,& (G_n,\Delta_n^{n-1},\Delta_{n-1}^n,\ldots,\Delta_n^1,\Delta_1^n) & G_i & = (G_n \cup \Delta_i^n) \setminus \Delta_n^i & \text{for }1 \leq i < n \\
+\mathcal{G}^{n-1}_n \da\,& (G_n,\Delta_n^{n-1},\Delta_{n-1}^n,\ldots,\Delta_2^1,\Delta_1^2) & G_i & = (G_{i+1} \cup \Delta_i^{i+1}) \setminus \Delta_{i+1}^i & \text{for }1 \leq i < n
+\end{align*}%
+}%
+%
+As can be seen for the definitions of $G_i$, these transformations are lossless: we can retrieve any version of the graph from any such transformation.
+The first two transformations start with the earliest version as a base.
+The first encodes deltas always with respect to the first version.
+The second encodes deltas with respect to the previous version.
+The latter two transformations start with the latest version.
+The third encodes deltas with respect to the latest version.
+The second encodes deltas with respect to the subsequent version.
+Letting $\mathcal{H} = (H_1,\ldots,H_n)$ such that $H_i = G_{n-i+1}$ ($1 \leq i \leq n$), i.e., such that $\mathcal{H}$ ``reverses''$\mathcal{G}$, we remark that $\mathcal{G}^n_1 = \mathcal{H}^1_n$ and $\mathcal{G}^n_{n-1} = \mathcal{H}^{n-1}_n$.
+Each such transformation can then be represented as a SPARQL dataset with $2(n-1)+1$ named graphs.
+
+In terms of space we expect $\mathcal{G}^n_{n-1}$ and $\mathcal{G}^{n-1}_n$ to be the most efficient as they always encode deltas from a neighbouring version.
+However, in terms of query rewriting, we expect $\mathcal{G}^n_{1}$ and $\mathcal{G}^{1}_n$ to be more efficient as they do not require a recursive construction of all intermediate versions towards the base version.
+In terms of indexing a new version, we expect $\mathcal{G}^{n}_1$ followed by $\mathcal{G}^{n}_{n-1}$ to be most efficient as they require only computing the most recent deltas; we expect $\mathcal{G}^{1}_n$, followed by $\mathcal{G}^{n-1}_{n}$, to be much more expensive, requiring a recompute of all deltas.
+On the other hand, $\mathcal{G}^{1}_n$ and $\mathcal{G}^{n-1}_{n}$ should be advantageous for queries over more recent versions, which we could reasonably expect to be the most frequent.\footnote{One may note that these four representations are analogous to differential backups, incremental backups, reverse-differential backups, and reverse-incremental backups.}
+
+\paragraph{Timestamp-Based (TB):} The intuition of the $TB$ representation is to associate each triple with the versions in which it is contained.
+Along these lines, we denote by $\mathcal{G}(s,p,o) \coloneqq \{ i \mid (s,p,o) \in G_i\text{ for } 1 \leq i \leq n \}$ the versions containing $(s,p,o)$.\footnote{The definition $\mathcal{G} : \I \times \I \times (\I \cup \L) \rightarrow 2^{\{1,\ldots,n\}}$~\cite{FernandezUPK19} is analogous to $\mathcal{G} = (G_1,\ldots,G_n)$.} Let $\mathcal{N} \coloneqq \{ N \mid \exists (s,p,o) \in G_i : \mathcal{G}(s,p,o) = N, 1 \leq i \leq n  \}$ denote the family of sets of versions associated with some triple in $(s,p,o)$.
+We can then represent the RDF archive with a named graph for each $N \in \mathcal{N}$.
+However, care must be taken as the number of named graphs could potentially reach $2^n$.
+Though this worst-case is unlikely to happen in practice (it would require triples often being intermittently added and removed across versions), a more practical option is to apply interval encoding and create a named graph for each interval~\cite{FernandezPU15}.
+More specifically, a triple $(s,p,o)$ is added to an interval $[i,j]$ (for $1 \leq i \leq j \leq n$) if and only if $(s,p,o) \in G_k$ for $i \leq k \leq j$ and either $i = 1$ or $(s,p,o) \notin G_{i-1}$ and $j = n$ or $(s,p,o) \notin G_{j+1}$; in simpler terms, $[i,j]$ is a maximal contiguous interval of versions in which $(s,p,o)$ appears.
+The upper bound is now $n(n+1)/2$ named graphs, but with the possibility of ``intermittent'' triples appearing in multiple named graphs for different intervals.
+Empty intervals can be omitted.
+
+In general, we expect the space to be similar to $\mathcal{G}^n_{n-1}$ and $\mathcal{G}^{n-1}_n$; in other words, quite good.
+In terms of query rewriting, we would expect performance to be quite good for a low number of versions and few intermittent triples as -- unlike for $CB$ representations -- negation is not required; we can rather take the union of interval graphs containing the version.
+However, as the number of versions $n$ grows, $O(n^2)$ interval graphs need to be unioned in the worst-case to materialise a particular version; by way of comparison, $CB$ representations require processing $O(1)$ (in the case of $\mathcal{G}^n_1$ and $\mathcal{G}^1_n$) or $O(n)$ (in the case of $\mathcal{G}^n_{n-1}$ and $\mathcal{G}^{n-1}_n$) named graphs to materialise a particular version.
+
+%There are a fours ways in which this can be done.
+The first option is to transform the RDF archive $\mathcal{G} = (G_1,\ldots,G_n)$ into a set of deltas relative to the initial version: $\mathcal{G}^1_n = (G_1,\Delta_1^2,\Delta_2^1,\ldots,\Delta_1^n,\Delta_n^1)$, resulting in $2(n-1)+1$ graphs.
+A particular version $G_i$ can then be reconstructed by applying $(G_1 \cup \Delta_i^1) \setminus \Delta_1^i$.
 

### 2020-09-09 query-rewrite.tex
added triple pattern rewriting for single version queries

Base pattern:
\begin{verbatim}
?s :p ?o
\end{verbatim}

\section{Single Version queries}
\subsection{Baseline}
The relevant version is select using a $\mathtt{FROM}$ or $\mathtt{GRAPH}$ statement.
\begin{verbatim}
FROM :i
\end{verbatim}
\vdots
\begin{verbatim}
?s :p ?o
\end{verbatim}
Alternatively:
\begin{verbatim}
GRAPH :i {?s :p ?o}
\end{verbatim}

\subsection{Intervals}
$\forall j \leq i, k \geq i$
\begin{verbatim}
FROM :j-k
\end{verbatim}
\vdots
\begin{verbatim}
?s :p ?o
\end{verbatim}
Alternatively $\mathtt{UNION}$ all :j-k where j $\leq$ i $\leq$ k in patterns such as:
\begin{verbatim}
GRAPH :j-k {?s :p ?o}
\end{verbatim}
\subsection{Absolute deltas}
The base version and ``positive'' deltas are combined as the default graph and the negative deltas are subtracted from the results.
\subsubsection{First version as base}
\begin{verbatim}
FROM :1
FROM :2-1
FROM NAMED :1-2
FROM :3-1
FROM NAMED :1-3
\end{verbatim}
\vdots
\begin{verbatim}
FROM :i-1
FROM NAMED 1-i
\end{verbatim}
\vdots
\begin{verbatim}
?s :p ?o
MINUS { GRAPH :1-2 {?s :p ?o} }
MINUS { GRAPH :1-3 {?s :p ?o} }
\end{verbatim}
\vdots
\begin{verbatim}
MINUS { GRAPH :1-i {?s :p ?o} }
\end{verbatim}

\subsubsection{Last version as base}
\begin{verbatim}
FROM :n
FROM :(n-1)-n
FROM NAMED :n-(n-1)
FROM :(n-2)-n
FROM NAMED :n-(n-2)
\end{verbatim}
\vdots
\begin{verbatim}
FROM :i-n
FROM NAMED n-i
\end{verbatim}
\vdots
\begin{verbatim}
?s :p ?o
MINUS { GRAPH :n-(n-1) {?s :p ?o} }
MINUS { GRAPH :n-(n-2) {?s :p ?o} }
\end{verbatim}
\vdots
\begin{verbatim}
MINUS { GRAPH :n-i {?s :p ?o} }
\end{verbatim}

\subsection{Sequential deltas}
The triple is searched in all deltas from the base up to the target. If the most recent delta is ``positive'', the match is preserved, otherwise it is discarded.
\subsubsection{First version as base}
\begin{verbatim}
{ SELECT  ?s ?o (MIN(?v0_) AS ?max_v0)
  WHERE
    { { { GRAPH ?g0
            { ?s  :p  ?o }
          ?g0  :direction  "forwards" .
          ?g0  :version  ?v0
        }
        BIND(str(?v0) AS ?v0_)
      }
      FILTER ( str(?v0) <= i )
    }
  GROUP BY ?s ?o
}
OPTIONAL
{ { SELECT  ?s ?o (MAX(?v1_) AS ?max_v1)
  WHERE
    { { { GRAPH ?g1
            { ?s  :p  ?o }
          ?g1  :direction  "backwards" .
          ?g1  :version  ?v1
        }
        BIND(str(?v1) AS ?v1_)
      }
      FILTER ( str(?v1) <= i )
    }
  GROUP BY ?s ?o
}
FILTER ( ( ! bound(?max_v1) ) || ( str(?max_v0) > str(?max_v1) ) )
}
\end{verbatim}
\subsubsection{Last version as base}
\begin{verbatim}
{ SELECT  ?s ?o (MIN(?v0_) AS ?max_v0)
  WHERE
    { { { GRAPH ?g0
            { ?s  <http://ex.org/p>  ?o }
          ?g0  :direction  "backwards" .
          ?g0  :version  ?v0
        }
        BIND(str(?v0) AS ?v0_)
      }
      FILTER ( str(?v0) >= i )
    }
  GROUP BY ?s ?o
}
OPTIONAL
{ { SELECT  ?s ?o (MIN(?v1_) AS ?max_v1)
  WHERE
    { { { GRAPH ?g1
            { ?s  :p  ?o }
          ?g1  :direction  "forwards" .
          ?g1  :version  ?v1
        }
        BIND(str(?v1) AS ?v1_)
      }
      FILTER ( str(?v1) >= i )
    }
  GROUP BY ?s ?o
}
FILTER ( ( ! bound(?max_v1) ) || ( str(?max_v0) < str(?max_v1) ) )
}
\end{verbatim}

\section{Version Delta queries}

### 2020-09-09 query-rewrite.tex
fixed some rewrittings and added version delta queries

@@
 \subsection{Intervals}
-$\forall j \leq i, k \geq i$
+All intervals containing the target version are combined as the default graph and the triple pattern is kept unmodified.
+
+$\forall$ j $\leq$ i, k $\geq$ i
 \begin{verbatim}
@@
 \subsection{Absolute deltas}
-The base version and ``positive'' deltas are combined as the default graph and the negative deltas are subtracted from the results.
+The base version and ``positive'' delta are combined as the default graph and the ``negative'' delta is subtracted from the results.
 \subsubsection{First version as base}
@@
 FROM :1
-FROM :2-1
-FROM NAMED :1-2
-FROM :3-1
-FROM NAMED :1-3
-\end{verbatim}
-\vdots
-\begin{verbatim}
 FROM :i-1
-FROM NAMED 1-i
+FROM NAMED :1-i
 \end{verbatim}
@@
 ?s :p ?o
-MINUS { GRAPH :1-2 {?s :p ?o} }
-MINUS { GRAPH :1-3 {?s :p ?o} }
-\end{verbatim}
-\vdots
-\begin{verbatim}
 MINUS { GRAPH :1-i {?s :p ?o} }
@@
 FROM :n
-FROM :(n-1)-n
-FROM NAMED :n-(n-1)
-FROM :(n-2)-n
-FROM NAMED :n-(n-2)
-\end{verbatim}
-\vdots
-\begin{verbatim}
 FROM :i-n
-FROM NAMED n-i
+FROM NAMED :n-i
 \end{verbatim}
@@
 ?s :p ?o
-MINUS { GRAPH :n-(n-1) {?s :p ?o} }
-MINUS { GRAPH :n-(n-2) {?s :p ?o} }
-\end{verbatim}
-\vdots
-\begin{verbatim}
 MINUS { GRAPH :n-i {?s :p ?o} }
@@
     { { { GRAPH ?g0
-            { ?s  <http://ex.org/p>  ?o }
+            { ?s  :p  ?o }
           ?g0  :direction  "backwards" .
@@
 \section{Version Delta queries}
+\subsection{Baseline}
+The target version is select using a $\mathtt{FROM}$ or $\mathtt{GRAPH}$ statement, while the preceding version can only be accessed using a $\mathtt{GRAPH}$ statement.
+\begin{verbatim}
+FROM :i
+FROM NAMED :(i-1)
+\end{verbatim}
+\vdots
+\begin{verbatim}
+?s :p ?o
+MINUS { GRAPH :(i-1) {?s :p ?o} }
+\end{verbatim}
+Alternatively:
+\begin{verbatim}
+GRAPH :i {?s :p ?o}
+MINUS { GRAPH :(i-1) {?s :p ?o} }
+\end{verbatim}
+
+\subsection{Intervals}
+All intervals containing the target version are combined as the default graph and the triple pattern is kept unmodified.
+All intervals containing the preceding version are accessed as named graphs and subtracted from the results.
+
+$\forall$ j $\leq$ i, k $\geq$ i; $\forall$ l $\leq$ (i-1), m $\geq$ (i-1)
+\begin{verbatim}
+FROM :j-k
+FROM NAMED :l-m
+\end{verbatim}
+\vdots
+\begin{verbatim}
+?s :p ?o
+MINUS { GRAPH :l-m {?s :p ?o} }
+\end{verbatim}
+Alternatively $\mathtt{UNION}$ all :j-k where j $\leq$ i $\leq$ k in patterns such as:
+\begin{verbatim}
+GRAPH :j-k {?s :p ?o}
+\end{verbatim}
+and then subtract the same patterns as above.
+\subsection{Absolute deltas}
+The base version and ``positive'' deltas are combined as the default graph and the negative deltas are subtracted from the results to build the target version.
+For the preceding version, the base and ``positive'' delta are combined with an $\mathtt{UNION}$ statement and the ``negative'' delta is substracted with a $\mathtt{MINUS}$ statement.
+\subsubsection{First version as base}
+\begin{verbatim}
+FROM :1
+FROM :i-1
+FROM NAMED :1-i
+FROM NAMED :1
+FROM NAMED :(i-1)-1
+FROM NAMED :1-(i-1)
+\end{verbatim}
+\vdots
+\begin{verbatim}
+{
+?s :p ?o
+MINUS { GRAPH :1-i {?s :p ?o} }
+}
+MINUS {
+{	{ GRAPH :1 {?s :p ?o} }
+	UNION
+	{ GRAPH :(i-1)-1 {?s :p ?o} }
+} MINUS { GRAPH :1-(i-1) {?s :p ?o} }
+}
+\end{verbatim}
+
+\subsubsection{Last version as base}
+\begin{verbatim}
+FROM :n
+FROM :i-n
+FROM NAMED :n-i
+FROM NAMED :n
+FROM NAMED :(i-1)-n
+FROM NAMED :n-(i-1)
+\end{verbatim}
+\vdots
+\begin{verbatim}
+{
+?s :p ?o
+MINUS { GRAPH :n-i {?s :p ?o} }
+}
+MINUS {
+{	{ GRAPH :n {?s :p ?o} }
+	UNION
+	{ GRAPH :(i-1)-n {?s :p ?o} }
+} MINUS { GRAPH :n-(i-1) {?s :p ?o} }
+}
+\end{verbatim}
+
+\subsection{Sequential deltas}
+For sequential deltas, Version Delta queries are handled as two separate queries. Each of these is rewritten as a Single Version query.
+\end{verbatim}

### 2020-09-09 paper.tex
minor edits
Experiments

@@
 \usepackage{etoolbox}
+
+\renewcommand{\UrlFont}{\ttfamily\scriptsize}
 
@@
 
+\newcommand{\ic}{\textsc{i}\xspace}
+\newcommand{\cbpd}{\ensuremath{\textsc{c}_d^{+}}\xspace}
+\newcommand{\cbpi}{\ensuremath{\textsc{c}_i^{+}}\xspace}
+\newcommand{\cbmd}{\ensuremath{\textsc{c}_d^{-}}\xspace}
+\newcommand{\cbmi}{\ensuremath{\textsc{c}_i^{-}}\xspace}
+\newcommand{\cbi}{\ensuremath{\textsc{c}_i}\xspace}
+\newcommand{\cbd}{\ensuremath{\textsc{c}_d}\xspace}
+\newcommand{\cbp}{\ensuremath{\textsc{c}^{+}}\xspace}
+\newcommand{\cbm}{\ensuremath{\textsc{c}^{-}}\xspace}
+\newcommand{\tb}{\textsc{t}\xspace}
+
 \paragraph{Independent Copies (IC):} Perhaps the most intuitive representation is to store each constituent graph of an RDF Archive $\mathcal{G} = (G_1,\ldots,G_n)$ as a SPARQL dataset $D = \{ (x_1,G_1), \ldots, (x_n,G_n)\}$, where $x_1,\ldots,x_n$ are IRIs that deterministically identify the version.
@@
 The third encodes deltas with respect to the latest version.
-The second encodes deltas with respect to the subsequent version.
+The forth encodes deltas with respect to the subsequent version.
 Letting $\mathcal{H} = (H_1,\ldots,H_n)$ such that $H_i = G_{n-i+1}$ ($1 \leq i \leq n$), i.e., such that $\mathcal{H}$ ``reverses''$\mathcal{G}$, we remark that $\mathcal{G}^n_1 = \mathcal{H}^1_n$ and $\mathcal{G}^n_{n-1} = \mathcal{H}^{n-1}_n$.
@@
 
+\begin{verbatim}
+?s :p ?o
+\end{verbatim}
+
 \section{Experiments}
+
+We now perform experiments to address the following research questions (Q1) Which of the six representations allow for better compression, more efficient indexing, and more efficient updates of a new version?
+(Q2) How do the query runtimes of compressed representations (\textit{CB}, \textit{TB}) compare with indexing complete versions (\textit{IC})?
+(Q3) Which representation works best overall?
+
+\paragraph{Setting} We address these research questions in the context of a Wikidata archive of 8 weekly truthy versions from 2017-08-09 with 1.506 billion triples, to 2017-09-27 with 1.924 billion triples.
+The RDF archive thus consists of 13.477 billion versioned triples.
+Each week 25--93 million triples are added, while 4--6 million triples are removed (an update counts as both an addition and a removal).
+We base our experiments on the example list of queries contributed by users and members of the Wikidata community, which features a wide variety of query operators, diverse terms, etc.\footnote{\url{https://www.wikidata.org/wiki/Wikidata:SPARQL_query_service/queries/examples}} We translate some of the Wikidata-specific features (e.g., the label service) to standard SPARQL.
+We further filter queries that feature federation to other endpoints, property paths, and qualifiers (not present in the truthy versions).
+The result is a test set of 146 SPARQL queries.
+We choose Virtuoso as our SPARQL implementation since it is widely used and has demonstrated good performance in prior benchmarks over Wikidata~\cite{HernandezHK15}.
+The runtimes presented are averaged over three runs.
+A timeout of 5 minutes was set per query.
+
+\paragraph{Indexing} We first look at the results of total index sizes for each representation.
+In Figure~\ref{fig:index} we show the index sizes (GB) for each representation, the time taken (min.) to bulk load all versions in the representation, and the time taken (min.) to update the archive indexing seven versions with the eighth version.
+We see that (as expected) \ic has the largest index sizes, followed by \cbpd and \cbmd.
+followed by \tb, followed by \cbpi and \cbmi with the smallest indexes.
+We see that for a bulk load, index times correlate with index size, with \ic being by far the slowest due to the greater volume of data to index.
+Finally, with respect to the version update, we again see a similar trend, except that (as expected) \cbmi is far slower than the other alternatives (even \ic) as the entire archive must be built from scratch relative to the most recent version, taking time comparable to a full bulk load.
+
+\paragraph{Single version queries} We apply the rewriting of our 146 SPARQL queries for each of our six representations in order to retrieve results for version 1, 5 and 8.
+We show the results as box-plots with log $y$-axis in Figure~\ref{fig:svq}.
+In general we note that median times were generally in the range of 100-1000~ms, though 1--20 queries timed-out in each experiment, affecting the mean (shown as a diamond).
+In terms of mean and median runtimes, \ic performs the best, with \cbpd, \cbmd and \tb also performing competitively across the different versions.
+Conversely, \cbpi and \cbmi perform quite poorly relative to the other options (except \cbmi in the case of version 8, which performs well in the median case).
+Contrasting the query times with index sizes, we note a clear time--space trade-off, where the largest index performs best, the smallest perform worst, and those with intermediate space perform in the middle of the pack.
+
+\paragraph{Delta version queries} Next we rewrite our 146 SPARQL queries in order to retrieve delta results between versions 1--2, 4--5 and 7--8 from each of our six representations.
+We again show the results as box-plots with log $y$-axis in Figure~\ref{fig:dvq}.
+When compared with single version queries, we see a considerable leap in time, where the median runtimes of even the best performing representations approach or exceed 1000 ms more frequently.
+This time we note that the best performance is offered by \tb, followed by \ic, \cbpi and \cbmi.
+Conversely,  \cbpd and \cbmd perform very poorly.
+In these experiments the incremental builds perform better than differential builds as the incremental deltas directly provide the difference between subsequent versions, providing an additional avenue for optimisation.
+
+\begin{figure}[t]
+\includegraphics[scale=0.8]{plots/indexsize}
+\includegraphics[scale=0.8]{plots/indextime}
+\includegraphics[scale=0.8]{plots/updatetime}
+\caption{Indexing details including size (left), bulk load (mid) and version update (right) \label{fig:index}}
+\end{figure}
+
+\begin{figure}[t]
+\includegraphics[scale=0.8]{plots/boxplt-svq}
+\caption{Single version query times for version 1 (left), 5 (mid) and 8 (right) \label{fig:svq}}
+\end{figure}
+
+\begin{figure}[t]
+\includegraphics[scale=0.8]{plots/boxplt-dvq}
+\caption{Delta version query times for version 1--2 (left), 4--5 (mid) and 7--8 (right) \label{fig:dvq}}
+\end{figure}
 

### 2020-09-09 paper.tex
first submitted version

--- 2020-09-08 paper.tex
@@
 \begin{abstract}
-Although many prominent Linked Datasets are in a state of constant evolution, historical data are rarely published in a standard way.
-Though a number of works have recently addressed the issue of querying and managing archives of versioned RDF data, most propose dedicated solutions, which incurs a high deployment cost.
-We rather explore solutions for representing archives of versioned RDF data using the SPARQL standard without modification, allowing existing implementations to be used.
-For querying, we consider two versioning primitives: viewing results from a particular version, or viewing results that change between two consecutive versions.
-We identify and compare several models by which historical versions of an RDF graph can be stored as named graphs, and discuss how base queries can be rewritten in each model per the two aforementioned primitives.
-We then present a performance comparison of these models and baseline (non-)versioned queries with respect to 23 weekly versions of Wikidata and 232 user-defined queries, using Virtuoso as a reference SPARQL implementation.
+We explore solutions for representing archives of versioned RDF data using the SPARQL standard and off-the-shelf engines.
+We consider six representations of RDF archives based on named graphs, and describe how input queries can be rewritten to return solutions for a particular version, or solutions that change between versions.
+We evaluate these alternatives over an archive of 8 weekly versions of Wikidata and 146 queries, using Virtuoso as the SPARQL engine.
 \end{abstract}
@@
 Within these works, a variety of specialised indexing techniques~\cite{Cerdeira-PenaFF16,BahriLA18,TaelmanSHMV19}, query languages~\cite{TappoletB09} and benchmarks~\cite{KotsevMPEFK16,FernandezUPK19} have been proposed, developed and evaluated.
-While these represent important advances, many such proposals are by their nature incompatible with the existing SPARQL infrastructure.
-Furthermore, benchmarks have mostly focused on synthetic data.
-Hence there is still a considerable gap between these research works and putting RDF archives into practice.
+While these represent important advances, many such works propose custom SPARQL extensions, engines, etc., creating an barrier for adoption.
 
@@
 
-In order to bridge this gap, we highlight that -- in theory at least -- there is no need for specialised indexes, query languages, etc., but rather that the types of versioned queries proposed in the literature~\cite{FernandezPU15} can be supported using off-the-shelf SPARQL engines that already enjoy years of development, optimisation, etc., as well as broad deployment on the Web.
-For example, SPARQL named graphs can be used to track different versions of individual graphs.
+In fact, the types of versioned queries proposed in the literature~\cite{FernandezPU15} can be supported using off-the-shelf SPARQL engines that already enjoy years of development, optimisation, and deployment.
+SPARQL named graphs can, for exampe, be used to track different versions of individual graphs.
 However, as Fernandez at al.~\cite{FernandezUPK19} note, the approach of using pure SPARQL would ``\textit{typically render rather inefficient SPARQL queries}''.
-This claim raises the research question: how inefficient will pure SPARQL be in this setting?
-Though Fernandez at al.~\cite{FernandezUPK19} do discuss how pure SPARQL could be used, to the best of our knowledge, this question has not been empirically explored.
-If a pure SPARQL solution could be found with relatively little overhead versus base queries (over a single current version) then the benefits would be significant in terms of being able to leverage existing SPARQL infrastructure -- i.e., languages, protocols, implementations -- for managing and querying RDF archives.
+This raises a question: how inefficient will such queries be?
+If a pure SPARQL solution could be found with reasonable performance, existing SPARQL infrastructure could be leveraged for RDF archives.
 
 In this paper, we present preliminary empirical results addressing this research question.
-Specifically we take the Virtuoso SPARQL engine~\cite{Erling12} and  evaluate five different representations of RDF archives for XX different weekly versions of truthy Wikidata dumps~\cite{VrandecicK14}.
-We then implement query rewriting mechanisms for each representation in order to support single-version queries and sequential single-delta queries that retrieve all solutions over an indicated version and all solutions that differ between two sequential versions, respectively.
-We then perform experiments to compare the sizes of indexes, the costs of indexing new versions, and the costs of query evaluation for base queries over the current version versus rewritten versioned queries.
+Specifically we look at six representations of RDF archives using named graphs and propose query rewriting mechanisms for them.
+We then evaluate and compare these representations for an RDF archive of 8 versions of Wikidata~\cite{VrandecicK14}.
+Our experiments compare the sizes of the indexes generated, the time taken for indexing, and the relative costs of query evaluation.
 
@@
 
-Various temporal extensions for RDF/SPARQL have been proposed in literature.
+%Various temporal extensions for RDF/SPARQL have been proposed in literature.
 Gutierrez et al.~\cite{GutierrezHV07} describe a framework that allows time representation in RDF using annotations.
@@
 Zimmerman et al.~\cite{ZimmermannLPS12} provide a general framework for annotating RDF that generalises temporal annotations; they further propose a query language called AnQL for querying annotated RDF.
+Proposed temporal extensions for SPARQL include $\tau$-SPARQL~\cite{TappoletB09}, T-SPARQL~\cite{Grandi10}, SPARQL-ST~\cite{PerryJS11}, SPARQL\textsuperscript{T}~\cite{ZanioloGACG18}, etc.
+Related to temporality, a number of systems support versioning for RDF, including SemVersion~\cite{VolkelG06}, POI~\cite{TzitzikasTA08}, x-RDF-3x~\cite{NeumannW10}, R43ples~\cite{GraubeHU14}, Dydra~\cite{AndersonB16} and Ostrich~\cite{TaelmanSV18}.
+
+Various temporal extensions for RDF have been proposed in literature based on annotations~\cite{GutierrezHV07,PuglieseUS08,ZimmermannLPS12}.
 Proposed temporal extensions for SPARQL include $\tau$-SPARQL~\cite{TappoletB09}, T-SPARQL~\cite{Grandi10}, SPARQL-ST~\cite{PerryJS11}, SPARQL\textsuperscript{T}~\cite{ZanioloGACG18}, etc.
@@
 
-In summary, the past years have seen a wide range of developments in terms of representations, languages, vocabularies, indexes, algorithms, etc., for managing and querying over temporal and/or versioned RDF data.
-However, to the best of our knowledge, most of these approaches propose specialised languages, implementations, etc., for these purposes.
-Such languages have the benefit of allowing to more concisely and intuitively express versioned queries, while such implementations can develop optimised techniques that leverage particular characteristics of versioned data.
+The past years have thus seen a wide range of developments in terms of representations, languages, vocabularies, indexes, algorithms, etc., for managing and querying over temporal and/or versioned RDF data.
+However, to the best of our knowledge, most of these approaches propose specialised languages, implementations, etc., for these purposes, creating an obstacle for adoption.
+%Such languages have the benefit of allowing to more concisely and intuitively express versioned queries, while such implementations can develop optimised techniques that leverage particular characteristics of versioned data.
 The downside of these specialised approaches is that they are often diverge from practice, requiring the use of research prototypes rather than established SPARQL implementations, non-standard languages, etc.
-A number of authors have acknowledged that one can manage and query RDF archives using vanilla SPARQL, though it may lead to prolix or inefficient queries~\cite{TappoletB09,FernandezUPK19}; however, we are not aware of any empirical results along these lines.
-Our work presents some preliminary results that aim to fill this gap and to better understand the costs and limitations of using vanilla SPARQL to host RDF archives.
+A number of authors note that one can manage and query RDF archives using vanilla SPARQL, though it may lead to prolix or inefficient queries~\cite{TappoletB09,FernandezUPK19}.
+This paper describes preliminary experiments to gain insights into how (in)efficient vanilla SPARQL and off-the-shelf engines are for hosting RDF archives.
 
@@
 
-We first establish some preliminaries and notation.
+We establish some preliminaries and notation.
 RDF triples are composed of three sets of terms: IRIs (\I), literals (\L) and blank nodes (\B).
-In this preliminary work, we do not consider blank nodes, whose special semantics complicates the detection of changes in RDF graphs~\cite{ZeginisTC11}.
+In this preliminary work, we do not consider blank nodes, whose special semantics complicates the detection of changes~\cite{ZeginisTC11}.
 Leaving aside blank nodes, an RDF triple $(s,p,o) \in \I \times \I \times (\I \cup \L)$ consists of a subject $s$, predicate $p$ and object $o$.
@@
 
-A triple pattern $(s,p,o) \in (\I \cup \V) \times (\I \cup \V) \times (\I \cup \L \cup \V)$ is an RDF triple that permits variables from a set $\V$ to appear in any position.
+A triple pattern $t \coloneqq (s,p,o) \in (\I \cup \V) \times (\I \cup \V) \times (\I \cup \L \cup \V)$ is an RDF triple that permits variables from $\V$ to appear in any position.
+We denote by $\vars{t}$ the variables of $t$.
 A solution is a partial mapping $\mu : \V \rightarrow \I \cup \L$.
@@
 We say that two solutions $\mu_1, \mu_2$ are compatible, denoted $\mu_1 \sim \mu_2$, if and only if $\mu_1(v) = \mu_2(v)$ for all $v \in \dom{\mu_1} \cap \dom{\mu_2}$.
-
-A basic graph pattern $B$ is a set of triple patterns.
-Let $\vars{B}$ denote the set of variables appearing in $B$ and let $\mu(B)$ denote the image of $B$ under $\mu$, i.e., the result of replacing all variables $v \in \dom{\mu} \cap \vars{B}$ with $\mu(v)$ in $B$.
-The evaluation of a basic graph pattern $B$ on an RDF graph $G$ is given as $B(G) \coloneqq \{ \mu \mid \mu(B) \subseteq G\text{ and }\dom{\mu} = \vars{B} \}$.
-
-SPARQL queries are based on basic graph patterns and a number of relational operators.
-Similar to Pérez et al.~\cite{PerezAG09}, we define an abstract syntax for a subset of SPARQL of pertinence to this paper as follows.
-A basic graph pattern $B$ is a graph pattern.
-Furthermore, if $P$ and $Q$ are graph patterns, and $V \subset \V$ is a set of variables, then $\sand{P}{Q}$, $\suni{P}{Q}$, $\sminus{P}{Q}$ and $\ssel{P}{V}$ are graph patterns.
-We define the semantics of these graph patterns in Table~\ref{tab:sparql}.
+We denote by $\mu(t)$ the image of $t$ under $\mu$, replacing each variable $v \in \dom{\mu} \cap \vars{t}$ by $\mu(v)$ in $t$.
+We denote by $t(G) \coloneqq \{ \mu \mid \mu(t) \subseteq G\text{ and }\dom{\mu} = \vars{t} \}$ the evaluation of $t$ over $G$.
 
 \begin{table}[t]
-{\scriptsize
+{\footnotesize
 \begin{align*}
@@
 %\sigma_R(M) \da\,& \{ \mu \in M \mid \mu \models R \}\\
-\ssel{P}{V}(G) \da &\, \pi_V(P(G)) & \pi_V(M) \da\,& \{ \mu' \mid \exists \mu \in M : \mu' \sim \mu, \dom{\mu'} = V \cap \dom{\mu} \}
+%\ssel{P}{V}(G) \da &\, \pi_V(P(G)) & \pi_V(M) \da\,& \{ \mu' \mid \exists \mu \in M : \mu' \sim \mu, \dom{\mu'} = V \cap \dom{\mu} \}
 \end{align*}}
-\caption{Semantics for SPARQL query operators where $M$, $M_1$ and $M_2$ denote sets of solutions, $V$ denotes a set of variables\label{tab:sparql}}
+\vspace{-1cm}
+\caption{Semantics for query operators where $M$, $M_1$ and $M_2$ denote sets of solutions\label{tab:sparql}}
 \end{table}
 
-In this paper, we rely not only on RDF graphs, but also on SPARQL datasets to represent RDF archives.
-A SPARQL dataset $D \coloneqq \{ G, (n_1,G_1), \ldots (n_m,G_m) \}$ consists of a default (RDF) graph $G$ and a set of named graphs, each being a pair $(n_i,G_i)$ where $n_i \in \I$, $G_i$ is an RDF graph, and $n_i \neq n_j$ for $1 \leq i \leq m$, $1 \leq j \leq m$, $i \neq j$.
-For simplicity, we will consider a SPARQL dataset $D$ to be a set of quads of the form $(G \times \{*\}) \cup (G_1 \times \{n_1\}) \cup \ldots \cup (G_m \times \{n_m\})$, where $* \notin \I \cup \L \cup \V$ is a special symbol denoting the default graph.\footnote{We thus assume a \textit{quad store}, and disallow empty named graphs.} Different named graphs can be queried using a \texttt{GRAPH} operator, creating quad patterns.
-A quad pattern $(s,p,o,g) \in (\I \cup \V) \times (\I \cup \V) \times (\I \cup \L \cup \V) \times (\I \cup \{ *\})$ extends a triple pattern with a fourth element that may be an IRI or *.
-A basic named graph pattern $C$ is then a set of quad patterns whose evaluation on an SPARQL dataset $D$ is given as $C(D) \coloneqq \{ \mu \mid \mu(C) \subseteq D\text{ and }\dom{\mu} = \vars{C} \}$.
-Following the SPARQL standard~\cite{sparql11}, we translate a triple pattern $(s,p,o)$ to a quad pattern $(s,p,o,*)$ such that it is evaluated only on the default graph.\footnote{In concrete SPARQL syntax, quad patterns are expressed with the $\texttt{GRAPH}$ keyword.} The semantics of the operators defined in Table~\ref{tab:sparql} then remain unchanged simply allowing $P$ and $Q$ to now represent basic named graph patterns, defined to be graph patterns.
+%A basic graph pattern $B$ is a set of triple patterns.
+%Let $\vars{B}$ denote the set of variables appearing in $B$ and let $\mu(B)$ denote the image of $B$ under $\mu$, i.e., the result of replacing all variables $v \in \dom{\mu} \cap \vars{B}$ with $\mu(v)$ in $B$.
+The evaluation of a basic graph pattern $B$ on an RDF graph $G$ is given as $B(G) \coloneqq \{ \mu \mid \mu(B) \subseteq G\text{ and }\dom{\mu} = \vars{B} \}$.
+SPARQL queries are based on triple patterns and a number of relational operators.
+Similar to Pérez et al.~\cite{PerezAG09}, we define an abstract syntax for a subset of SPARQL of pertinence to this paper as follows.
+A triple pattern $t$ is a graph pattern.
+Furthermore, if $P$ and $Q$ are graph patterns, and $V \subset \V$ is a set of variables, then $\sand{P}{Q}$, $\suni{P}{Q}$ and $\sminus{P}{Q}$ are graph patterns.
+We define the semantics of these graph patterns in Table~\ref{tab:sparql}.\footnote{A basic graph pattern $\{ t_1, \ldots t_n\}$ can be represented as $\sand{\sand{t_1}{\ldots}}{t_n}$.}
+
+In this paper, we rely on SPARQL datasets to represent RDF archives.
+A SPARQL dataset $D \coloneqq \{ G, (x_1,G_1), \ldots (x_n,G_n) \}$ consists of a default (RDF) graph $G$ and a set of named graphs of the form $(x_i,G_i)$ where $x_i \in \I$, $G_i$ is an RDF graph, and $x_i \neq x_j$ for $1 \leq i \leq m$, $1 \leq j \leq m$, $i \neq j$.
+We may represent $D$ as a set of quads of the form $(G \times \{*\}) \cup (G_1 \times \{x_1\}) \cup \ldots \cup (G_n \times \{x_n\})$, where $* \notin \I \cup \L \cup \V$ is a special symbol denoting the default graph.\footnote{We thus assume a \textit{quad store}, and disallow empty named graphs.} Different named graphs can be queried using a \texttt{GRAPH} operator, creating quad patterns.
+A quad pattern $q = (s,p,o,g) \in (\I \cup \V) \times (\I \cup \V) \times (\I \cup \L \cup \V) \times (\I \cup \{ *\})$ extends a triple pattern with a fourth element that may be an IRI or *.
+Its evaluation is analogous to that of a triple pattern: $q(G) \coloneqq \{ \mu \mid \mu(q) \subseteq G\text{ and }\dom{\mu} = \vars{q} \}$.
+Following the SPARQL standard~\cite{sparql11}, we translate a triple pattern $(s,p,o)$ to a quad pattern $(s,p,o,*)$ evaluated only on the default graph.
+The semantics of the operators defined in Table~\ref{tab:sparql} then remain unchanged simply allowing $P$ and $Q$ to now also represent quad patterns, considered to be (named) graph patterns.
 
@@
 Likewise if $P$ is a graph pattern, and $M$ and $N$ are sets of IRIs, then $\sfrom{P}{M}{N}$ is also a query.
-The evaluation of $\sfrom{P}{M}{N}$ on a dataset $D$ is defined as $\sfrom{P}{M}{N}(D) \coloneqq P(D_{M,N})$, where the query dataset $D_{M,N} \coloneqq (\cup_{m \in M} D(m) \times \{ *\}) \cup  (\cup_{n \in N} D(n) \times n)$ is formed from a default graph containing the union\footnote{Since we do not allow blank nodes, a union or RDF merge is equivalent.} of graphs with a name $m \in M$, and all the named graphs in $D$ with a graph named $n \in N$.
+If $(x_i,G_i) \in D$ let $D(x_i) = G_i$; otherwise if no graph is named $x_i$ in $D$, let $D(x_i) = \{\}$.
+The evaluation of $\sfrom{P}{M}{N}$ on a dataset $D$ is defined as $\sfrom{P}{M}{N}(D) \coloneqq P(D_{M,N})$, where the query dataset $D_{M,N} \coloneqq (\cup_{m \in M} D(m) \times \{ *\}) \cup  (\cup_{n \in N} D(n) \times n)$ is formed from a default graph containing the union\footnote{Since we do not allow blank nodes, a union or RDF merge is equivalent.} of graphs with a name $m \in M$, and all named graphs in $D$ with a name $n \in N$.
 We will be able to use these operators to define query datasets that capture specific versions in an RDF archive.
@@
 
-\paragraph{Independent Copies (IC):} Perhaps the most intuitive representation is to store each constituent graph of an RDF Archive $\mathcal{G} = (G_1,\ldots,G_n)$ as a SPARQL dataset $D = \{ (x_1,G_1), \ldots, (x_n,G_n)\}$, where $x_1,\ldots,x_n$ are IRIs that deterministically identify the version.
-This will result in relatively simple (and thus likely efficient) rewritten queries, but can be expected to occupy a lot of space, particularly in cases where a small proportion of triples change from version to version.
-We thus consider the $IC$ representation to be a baseline for comparison.
+\newcommand{\ic}{\textsc{i}\xspace}
+\newcommand{\cbpd}{\ensuremath{\textsc{c}_d^{+}}\xspace}
+\newcommand{\cbpi}{\ensuremath{\textsc{c}_i^{+}}\xspace}
+\newcommand{\cbmd}{\ensuremath{\textsc{c}_d^{-}}\xspace}
+\newcommand{\cbmi}{\ensuremath{\textsc{c}_i^{-}}\xspace}
+\newcommand{\cbi}{\ensuremath{\textsc{c}_i}\xspace}
+\newcommand{\cbd}{\ensuremath{\textsc{c}_d}\xspace}
+\newcommand{\cbp}{\ensuremath{\textsc{c}^{+}}\xspace}
+\newcommand{\cbm}{\ensuremath{\textsc{c}^{-}}\xspace}
+\newcommand{\tb}{\textsc{t}\xspace}
 
-\paragraph{Change-Based (CB):} The core idea of $CB$ representations is to store only triples that change from a given reference version.
+\paragraph{Independent Copies (IC):} A natural representation is to store an RDF Archive $\mathcal{G} = (G_1,\ldots,G_n)$ as a SPARQL dataset $D = \{ (x_1,G_1), \ldots, (x_n,G_n)\}$, where $x_1,\ldots,x_n$ are IRIs that identify the version.
+This will result in relatively simple (and thus likely efficient) rewritten queries, but can be expected to occupy a lot of space, particularly where few triples change from version to version.
+%We thus consider the IC representation to be a baseline for comparison.
+
+\paragraph{Change-Based (CB):} The core idea of CB representations is to store only triples that change from a given reference version.
 Along these lines, in the following we denote by $\Delta_{i}^j \coloneqq G_i \setminus G_j$ the triples in version $i$ not in version $j$.
-We then consider four $CB$ representations based on the following transformations of an RDF archive $\mathcal{G} = (G_1,\ldots,G_n)$:
+We consider four CB representations based on four transformations of $\mathcal{G} = (G_1,\ldots,G_n)$:
 %
@@
 The third encodes deltas with respect to the latest version.
-The second encodes deltas with respect to the subsequent version.
-Letting $\mathcal{H} = (H_1,\ldots,H_n)$ such that $H_i = G_{n-i+1}$ ($1 \leq i \leq n$), i.e., such that $\mathcal{H}$ ``reverses''$\mathcal{G}$, we remark that $\mathcal{G}^n_1 = \mathcal{H}^1_n$ and $\mathcal{G}^n_{n-1} = \mathcal{H}^{n-1}_n$.
+The forth encodes deltas with respect to the subsequent version.
+Letting $\mathcal{H} = (H_1,\ldots,H_n)$ such that $H_i = G_{n-i+1}$ ($1 \leq i \leq n$), i.e., such that $\mathcal{H}$ ``reverses'' $\mathcal{G}$, we remark that $\mathcal{G}^n_1 = \mathcal{H}^1_n$ and $\mathcal{G}^n_{n-1} = \mathcal{H}^{n-1}_n$.
 Each such transformation can then be represented as a SPARQL dataset with $2(n-1)+1$ named graphs.
@@
 In terms of indexing a new version, we expect $\mathcal{G}^{n}_1$ followed by $\mathcal{G}^{n}_{n-1}$ to be most efficient as they require only computing the most recent deltas; we expect $\mathcal{G}^{1}_n$, followed by $\mathcal{G}^{n-1}_{n}$, to be much more expensive, requiring a recompute of all deltas.
-On the other hand, $\mathcal{G}^{1}_n$ and $\mathcal{G}^{n-1}_{n}$ should be advantageous for queries over more recent versions, which we could reasonably expect to be the most frequent.\footnote{One may note that these four representations are analogous to differential backups, incremental backups, reverse-differential backups, and reverse-incremental backups.}
+On the other hand, $\mathcal{G}^{1}_n$ and $\mathcal{G}^{n-1}_{n}$ should be advantageous for queries over more recent versions, which we could reasonably expect to be the most frequent.
 
-\paragraph{Timestamp-Based (TB):} The intuition of the $TB$ representation is to associate each triple with the versions in which it is contained.
+These four representations are analogous to differential backups, incremental backups, reverse-differential backups, and reverse-incremental backups.
+
+\paragraph{Timestamp-Based (TB):} The intuition of the TB representation is to associate each triple with the versions in which it is contained.
 Along these lines, we denote by $\mathcal{G}(s,p,o) \coloneqq \{ i \mid (s,p,o) \in G_i\text{ for } 1 \leq i \leq n \}$ the versions containing $(s,p,o)$.\footnote{The definition $\mathcal{G} : \I \times \I \times (\I \cup \L) \rightarrow 2^{\{1,\ldots,n\}}$~\cite{FernandezUPK19} is analogous to $\mathcal{G} = (G_1,\ldots,G_n)$.} Let $\mathcal{N} \coloneqq \{ N \mid \exists (s,p,o) \in G_i : \mathcal{G}(s,p,o) = N, 1 \leq i \leq n  \}$ denote the family of sets of versions associated with some triple in $(s,p,o)$.
 We can then represent the RDF archive with a named graph for each $N \in \mathcal{N}$.
-However, care must be taken as the number of named graphs could potentially reach $2^n$.
-Though this worst-case is unlikely to happen in practice (it would require triples often being intermittently added and removed across versions), a more practical option is to apply interval encoding and create a named graph for each interval~\cite{FernandezPU15}.
-More specifically, a triple $(s,p,o)$ is added to an interval $[i,j]$ (for $1 \leq i \leq j \leq n$) if and only if $(s,p,o) \in G_k$ for $i \leq k \leq j$ and either $i = 1$ or $(s,p,o) \notin G_{i-1}$ and $j = n$ or $(s,p,o) \notin G_{j+1}$; in simpler terms, $[i,j]$ is a maximal contiguous interval of versions in which $(s,p,o)$ appears.
-The upper bound is now $n(n+1)/2$ named graphs, but with the possibility of ``intermittent'' triples appearing in multiple named graphs for different intervals.
-Empty intervals can be omitted.
+However, care must be taken as the number of named graphs could potentially reach $2^n$ (or the total number of unique triples in $\mathcal{G}$).
+A more practical option is create a named graph for intervals~\cite{FernandezPU15}.
+More specifically, a triple $(s,p,o)$ is added to an interval $[i,j]$ (for $1 \leq i \leq j \leq n$) if and only if $(s,p,o) \in G_k$ for $i \leq k \leq j$ and either $i = 1$ or $(s,p,o) \notin G_{i-1}$ and $j = n$ or $(s,p,o) \notin G_{j+1}$; in simpler terms, $[i,j]$ is a maximal contiguous interval of versions in which $(s,p,o)$ appears (omitting empty intervals).
+The upper bound is now $n(n+1)/2$ named graphs, but with the possibility of triples appearing in multiple named graphs for different intervals.
 
 In general, we expect the space to be similar to $\mathcal{G}^n_{n-1}$ and $\mathcal{G}^{n-1}_n$; in other words, quite good.
-In terms of query rewriting, we would expect performance to be quite good for a low number of versions and few intermittent triples as -- unlike for $CB$ representations -- negation is not required; we can rather take the union of interval graphs containing the version.
-However, as the number of versions $n$ grows, $O(n^2)$ interval graphs need to be unioned in the worst-case to materialise a particular version; by way of comparison, $CB$ representations require processing $O(1)$ (in the case of $\mathcal{G}^n_1$ and $\mathcal{G}^1_n$) or $O(n)$ (in the case of $\mathcal{G}^n_{n-1}$ and $\mathcal{G}^{n-1}_n$) named graphs to materialise a particular version.
+In terms of query rewriting, we would expect performance to be quite good for a low number of versions and few intermittent triples as -- unlike for CB representations -- negation is not required; we can rather take the union of interval graphs containing the version.
+However, as the number of versions $n$ grows, $O(n^2)$ interval graphs need to be unioned in the worst-case to materialise a particular version; by way of comparison, CB representations require processing $O(1)$ (in the case of $\mathcal{G}^n_1$ and $\mathcal{G}^1_n$) or $O(n)$ (in the case of $\mathcal{G}^n_{n-1}$ and $\mathcal{G}^{n-1}_n$) named graphs to materialise a particular version.
+
+\paragraph{Notation:} We denote IC by $\ic$; incremental, differential, reverse-incremental, reverse-differential CB by $\cbpd$, $\cbpi$, $\cbmd$, $\cbmi$, resp.; and interval TB by $\tb$.
 
@@
 
+Given a SPARQL query $Q$ defined for an RDF graph $G$, we now describe rewritings of $Q$ to generate solutions for different versions.
+We first focus on rewritings of triple patterns and then generalise.
+We do not extend the SPARQL syntax, but rather assume that version parameters are defined in (e.g.) the HTTP request.
+
+\subsection{Single-Version Queries} A single version query returns $Q(G_v)$ for a specified version $v$.
+Our overall strategy is to use \texttt{FROM} to construct the graph of the version where possible, as is the case for all versions in \ic and \tb; for $G_1$ in \cbpd, \cbpi; and for $G_n$ in \cbmd, \cbmi.
+Otherwise we rewrite each individual triple pattern appearing in $Q$ in order to ensure that it generates the same solutions as it would if evaluated over the graph of the selected version.
+We now provide more details for each representation
+
+\paragraph{Independent Copies (IC)} For $\ic$ we rewrite $Q$ to \sfrom{Q}{\{\texttt{v}\}}{\{\}}, where \texttt{v} is the IRI that names the graph for version $v$
+
+\paragraph{Change-Based (CB)} Recalling the observation that forwards and reverse CB representations are analogous, for brevity we define the rewriting for the forwards direction (\cbpd, \cbpi) only; the reverse direction (\cbmd, \cbmi) follows naturally.
+For the differential representation, we load the base version and the positive delta into the default graph and, for each triple pattern, we subtract the negative delta which is queried using a quad pattern.
+Thus for $\cbpd$ and $\cbmd$ we rewrite $Q$ to  \sfrom{Q'}{\{\texttt{1},\texttt{v1}\}}{\{\texttt{1v}\}}, where \texttt{1}, \texttt{v1}, \texttt{1v} indicate the names of $G_1$, $\Delta_v^1$ and $\Delta_1^v$, respectively; and $Q'$ replaces each triple pattern $(s,p,o) \in Q$ with the named graph pattern $\sminus{(s,p,o,*)}{(s,p,o,\texttt{1v})}$.
+
+Unfortunately the incremental rewriting is more complex.
+A first idea would be to take the union of $G_1$ and all positive deltas $\Delta_2^1, \ldots, \Delta_v^{v-1}$ and then subtract the (union of the) negative deltas $\Delta_1^2, \ldots, \Delta_{v-1}^{v}$; unfortunately this would overlook triples that were removed from a version $1 \leq u < v$ but were added back in a later version $u < u' \leq v$ (and were not removed again in a version $u' < u'' \leq v$).
+Hence a recursive rewriting appears to be necessary.
+Let $Q_1 \coloneqq \sfrom{Q}{\{\texttt{1}\}}{\{\}}$; then $Q_2 \coloneqq \sfrom{Q'_1}{\{\texttt{1}\}}{\{\texttt{12},\texttt{21}\}}$, where $Q_1$ is the result of replacing each triple pattern $(s,p,o)$ in $Q_1$ by the named graph pattern $P_2 \coloneqq \sminus{\suni{(s,p,o,*)}{(s,p,o,\texttt{21})}}{(s,p,o,\texttt{12})}$.
+We can then apply this rewriting recursively: $Q_i \coloneqq \sfrom{Q_{i-1}'}{\{\texttt{1}\}}{\{\texttt{12},\ldots,\texttt{(i-1)i},\texttt{21},\ldots,\texttt{i(i-1)}\}}$, where $Q_{i-1}'$ replaces each named graph pattern $P_{i-1}$ appearing in $Q_{i-1}$ with the recursive pattern $P_i \coloneqq \sminus{\suni{P_{i-1}}{(s,p,o,\texttt{i(i-1)})}}{(s,p,o,\texttt{(i-1)i})}$.
+
+This rewriting leads to complex queries.
+We thus optimise using additional features of SPARQL.
+To sketch the idea, our goal is then to make sure that each triple pattern only matches triples that appear in a base version and were not removed, or that appear in a positive delta $\Delta_{j}^{i}$ such that $1 < i < j \leq v$ and do not appear in a (later) negative delta $\Delta_{k}^{l}$ for $j \leq k < l \leq v$.
+In practice, for each delta $\Delta_{j}^{i}$ stored as a named graph $\texttt{ji}$, in offline processing, we index meta-data of the form $(\texttt{ji},\texttt{ver},j,\texttt{\$})$, and $(\texttt{ji},\texttt{type},\texttt{pos},\texttt{\$})$ in the case that $i < j$ or $(\texttt{ij},\texttt{type},\texttt{neg},\texttt{\$})$ in the case that $j < i$ ($\texttt{\$} \in \I$ is a reserved name for the meta-data graph).
+We can then check the aforementioned condition by using aggregation to find the maximum version of a negative delta less than or equals $v$ in which the triple pattern matches, then filtering the base version or any positive delta earlier than this maximum version.
+While the resulting query is still quite complex, we found it to be more practical than the recursive rewriting.
+
+%\subsection{Absolute deltas}
+%
+%\subsubsection{First version as base}
+%\begin{verbatim}
+%FROM :1
+%FROM :i-1
+%FROM NAMED :1-i
+%\end{verbatim}
+%\vdots
+%\begin{verbatim}
+%?s :p ?o
+%MINUS { GRAPH :1-i {?s :p ?o} }
+%\end{verbatim}
+%
+%\subsubsection{Last version as base}
+%\begin{verbatim}
+%FROM :n
+%FROM :i-n
+%FROM NAMED :n-i
+%\end{verbatim}
+%\vdots
+%\begin{verbatim}
+%?s :p ?o
+%MINUS { GRAPH :n-i {?s :p ?o} }
+%\end{verbatim}
+%
+%\subsection{Sequential deltas}
+%The triple is searched in all deltas from the base up to the target. If the most recent delta is ``positive'', the match is preserved, otherwise it is discarded.
+%\subsubsection{First version as base}
+%\begin{verbatim}
+%{ SELECT  ?s ?o (MIN(?v0_) AS ?max_v0)
+%  WHERE
+%    { { { GRAPH ?g0
+%            { ?s  :p  ?o }
+%          ?g0  :direction  "forwards" .
+%          ?g0  :version  ?v0
+%        }
+%        BIND(str(?v0) AS ?v0_)
+%      }
+%      FILTER ( str(?v0) <= i )
+%    }
+%  GROUP BY ?s ?o
+%}
+%OPTIONAL
+%{ { SELECT  ?s ?o (MAX(?v1_) AS ?max_v1)
+%  WHERE
+%    { { { GRAPH ?g1
+%            { ?s  :p  ?o }
+%          ?g1  :direction  "backwards" .
+%          ?g1  :version  ?v1
+%        }
+%        BIND(str(?v1) AS ?v1_)
+%      }
+%      FILTER ( str(?v1) <= i )
+%    }
+%  GROUP BY ?s ?o
+%}
+%FILTER ( ( ! bound(?max_v1) ) || ( str(?max_v0) > str(?max_v1) ) )
+%}
+%\end{verbatim}
+%\subsubsection{Last version as base}
+%\begin{verbatim}
+%{ SELECT  ?s ?o (MIN(?v0_) AS ?max_v0)
+%  WHERE
+%    { { { GRAPH ?g0
+%            { ?s  :p  ?o }
+%          ?g0  :direction  "backwards" .
+%          ?g0  :version  ?v0
+%        }
+%        BIND(str(?v0) AS ?v0_)
+%      }
+%      FILTER ( str(?v0) >= i )
+%    }
+%  GROUP BY ?s ?o
+%}
+%OPTIONAL
+%{ { SELECT  ?s ?o (MIN(?v1_) AS ?max_v1)
+%  WHERE
+%    { { { GRAPH ?g1
+%            { ?s  :p  ?o }
+%          ?g1  :direction  "forwards" .
+%          ?g1  :version  ?v1
+%        }
+%        BIND(str(?v1) AS ?v1_)
+%      }
+%      FILTER ( str(?v1) >= i )
+%    }
+%  GROUP BY ?s ?o
+%}
+%FILTER ( ( ! bound(?max_v1) ) || ( str(?max_v0) < str(?max_v1) ) )
+%}
+%\end{verbatim}
+%
+%
+%Base pattern:
+%\begin{verbatim}
+%?s :p ?o
+%\end{verbatim}
+%
+%
+%\subsection{Baseline}
+%The relevant version is select using a $\mathtt{FROM}$ or $\mathtt{GRAPH}$ statement.
+%\begin{verbatim}
+%FROM :i
+%\end{verbatim}
+%\vdots
+%\begin{verbatim}
+%?s :p ?o
+%\end{verbatim}
+%Alternatively:
+%\begin{verbatim}
+%GRAPH :i {?s :p ?o}
+%\end{verbatim}
+
+\paragraph{Timestamp-Based (TB)} Let \texttt{i:j} denote the name of the graph for the interval of versions $[i,j]$.
+We rewrite $Q$ to the query $\sfrom{Q}{\{I\}}{\{\}}$, where $I$ is the set of IRIs naming intervals in which $v$ is contained; formally:  $I \coloneqq \{ \texttt{i:j} \mid i \leq v \leq j \}$.
+
+\subsection{Delta-Version Queries}
+
+Given a query $Q$, a control version $u$ and a target version $v$, delta-version queries give solutions in $Q(G_v) \setminus Q(G_u)$.
+The general strategy for rewriting is to construct a query $\sminus{Q_v}{Q_u}$ where $Q_v$ and $Q_u$ are the respective single-version queries.
+While we continue to use the default graph as before for $Q_v$, we must rather use named graphs to manage the positive elements of $Q_u$ (replacing \texttt{FROM} with \texttt{FROM NAMED}/\texttt{GRAPH}).
+As an example, for $\ic$, we rewrite $Q$ to \sfrom{Q'}{\{\texttt{v}\}}{\{\texttt{u}\}}, where $Q'$ replaces each triple pattern $(s,p,o)$ in $Q$ with $\sminus{(s,p,o,*)}{(s,p,o,\texttt{u})}$.
+
+\subsection{Other SPARQL Features}
+
+In the SPARQL algebra, only the evaluation of triple patterns and path patterns directly accept the graph as input.
+Hence, given a SPARQL query $Q$ (over a default graph), if we can individually rewrite each triple pattern and path pattern within $Q$ to generate solutions for $G_v$, then the rewritten query $Q'$ will generate precisely the solutions for $G_v$.
+We previously described this process for triple patterns.
+However, a complication arises because in SPARQL we can only evaluate paths on the default graph or the union of all named graphs in the query dataset: we cannot express paths that span \texttt{GRAPH} clauses.
+Hence while we can straightforwardly rewrite path patterns for $\ic$, $\cbpd$, $\cbmd$ and $\tb$, we are not aware of such a method for $\cbpi$ and $\cbpd$ due to how their queries interleave \texttt{GRAPH} clauses.
+
+%\paragraph{Independent Copies (IC)} For $\ic$, we rewrite $Q$ to \sfrom{Q'}{\{\texttt{v}\}}{\{\texttt{u}\}}, where $Q'$ replaces each triple pattern $(s,p,o)$ in $Q$ with $\sminus{(s,p,o,*)}{(s,p,o,\texttt{u})}$.
+%
+%\paragraph{Change-Based (CB)} For $\cbpd$, we rewrite $Q$ to \sfrom{Q'}{\{\texttt{1},\texttt{v1}\}}{\{\texttt{1v},\texttt{1},\texttt{u1},\texttt{1u}\}}, where $Q'$ replaces each triple pattern $(s,p,o)$ in $Q$ with the following pattern:
+%%
+%\[\sminus{\sminus{(s,p,o,*)}{(s,p,o,\!\texttt{1v})}}{\sminus{\suni{(s,p,o,\!\texttt{1})}{(s,p,o,\!\texttt{u1})}}{(s,p,o,\!\texttt{1u})}} \]
+%%
+%\noindent The initial $\sminus{(s,p,o,*)}{(s,p,o,\texttt{1v})}$ computes the results for $Q(G_v)$, from which the results for $Q(G_u)$ are subtracted.
+
+%
+%
+%%The target version is select using a $\mathtt{FROM}$ or $\mathtt{GRAPH}$ statement, while the preceding version can only be accessed using a $\mathtt{GRAPH}$ statement.
+%%\begin{verbatim}
+%%FROM :i
+%%FROM NAMED :(i-1)
+%%\end{verbatim}
+%%\vdots
+%%\begin{verbatim}
+%%?s :p ?o
+%%MINUS { GRAPH :(i-1) {?s :p ?o} }
+%%\end{verbatim}
+%%Alternatively:
+%%\begin{verbatim}
+%%GRAPH :i {?s :p ?o}
+%%MINUS { GRAPH :(i-1) {?s :p ?o} }
+%%\end{verbatim}
+%
+%\subsection{Absolute deltas}
+%The base version and ``positive'' deltas are combined as the default graph and the negative deltas are subtracted from the results to build the target version.
+For the preceding version, the base and ``positive'' delta are combined with an $\mathtt{UNION}$ statement and the ``negative'' delta is substracted with a $\mathtt{MINUS}$ statement.
+%\subsubsection{First version as base}
+%\begin{verbatim}
+%FROM :1
+%FROM :i-1
+%FROM NAMED :1-i
+%FROM NAMED :1
+%FROM NAMED :(i-1)-1
+%FROM NAMED :1-(i-1)
+%\end{verbatim}
+%\vdots
+%\begin{verbatim}
+%{
+%?s :p ?o
+%MINUS { GRAPH :1-i {?s :p ?o} }
+%}
+%MINUS {
+%{	{ GRAPH :1 {?s :p ?o} }
+%	UNION
+%	{ GRAPH :(i-1)-1 {?s :p ?o} }
+%} MINUS { GRAPH :1-(i-1) {?s :p ?o} }
+%}
+%\end{verbatim}
+%
+%\subsubsection{Last version as base}
+%\begin{verbatim}
+%FROM :n
+%FROM :i-n
+%FROM NAMED :n-i
+%FROM NAMED :n
+%FROM NAMED :(i-1)-n
+%FROM NAMED :n-(i-1)
+%\end{verbatim}
+%\vdots
+%\begin{verbatim}
+%{
+%?s :p ?o
+%MINUS { GRAPH :n-i {?s :p ?o} }
+%}
+%MINUS {
+%{	{ GRAPH :n {?s :p ?o} }
+%	UNION
+%	{ GRAPH :(i-1)-n {?s :p ?o} }
+%} MINUS { GRAPH :n-(i-1) {?s :p ?o} }
+%}
+%\end{verbatim}
+
+%\subsection{Intervals}
+%All intervals containing the target version are combined as the default graph and the triple pattern is kept unmodified.
+All intervals containing the preceding version are accessed as named graphs and subtracted from the results.
+%
+%
+%$\forall$ j $\leq$ i, k $\geq$ i; $\forall$ l $\leq$ (i-1), m $\geq$ (i-1)
+%\begin{verbatim}
+%FROM :j-k
+%FROM NAMED :l-m
+%\end{verbatim}
+%\vdots
+%\begin{verbatim}
+%?s :p ?o
+%MINUS { GRAPH :l-m {?s :p ?o} }
+%\end{verbatim}
+%Alternatively $\mathtt{UNION}$ all :j-k where j $\leq$ i $\leq$ k in patterns such as:
+%\begin{verbatim}
+%GRAPH :j-k {?s :p ?o}
+%\end{verbatim}
+%and then subtract the same patterns as above.
+
 \section{Experiments}
 
+We now perform experiments to address the following three research questions (Q1) Which of the six representations allow for better compression, more efficient indexing, and more efficient updates of a new version?
+(Q2) How do the query runtimes of compressed representations (\textit{CB}, \textit{TB}) compare with indexing complete versions (\textit{IC})?
+(Q3) Which representation works best overall?
+
+\paragraph{Setting} We address these questions for a Wikidata archive of 8 weekly truthy versions from 2017-08-09 with 1.506 billion triples, to 2017-09-27 with 1.924 billion triples.
+The RDF archive consists of 13.477 billion triple--version pairs.
+Each week 25--93 million triples are added, while 4--6 million triples are removed.
+We take Wikidata's example queries, defined by users\footnote{\url{https://www.wikidata.org/wiki/Wikidata:SPARQL_query_service/queries/examples}}, and translate Wikidata-specific features (e.g., the label service) to standard SPARQL.
+We further filter queries that feature federation to other endpoints, property paths (not supported by all representations), and qualifiers (not in truthy versions).
+The result is a test set of 146 SPARQL queries.
+We take Virtuoso as our SPARQL implementation.
+Runtimes are averaged over three runs.
+Query timeouts were set to 5 minutes.
+The machine used has 120GB of RAM and a standard SATA hard-disk.
+
+\paragraph{Indexing} We first look at the results of total index sizes for each representation.
+In Figure~\ref{fig:index} we show the index sizes (GB) for each representation, the time taken (min.) to bulk load all versions in the representation, and the time taken (min.) to update a seven-version archive with the eighth version.
+We see that \ic has the largest index sizes, followed by \cbpd and \cbmd, then \tb, and finally \cbpi and \cbmi.
+The bulk load times correlate with index size, with \ic (thus) being by far the slowest.
+We see a similar trend in version updates, except that \cbmi is far slower than the other alternatives (even \ic) as the entire archive must be built from scratch.
+
+\paragraph{Single version queries} We apply the rewriting of our 146 SPARQL queries for each of our six representations in order to retrieve results for version 1, 5 and 8.
+We show the results as box-plots with log $y$-axis in Figure~\ref{fig:svq}.
+In general we note that median times were generally in the range of 100-1000~ms, though 1--20 queries timed out in each experiment, affecting the mean (shown as a diamond).
+In terms of mean and median runtimes, \ic performs the best, with \cbpd, \cbmd and \tb also performing competitively across the different versions.
+Conversely, \cbpi and \cbmi perform poorly relative to the other options (except \cbmi in the case of version 8).
+Contrasting the query times with index sizes, we note a clear time--space trade-off, where the largest index performs best, the smallest perform worst, and those with intermediate space perform in the middle of the pack.
+
+\paragraph{Delta version queries} Next we rewrite our 146 SPARQL queries in order to retrieve delta results between versions 1--2, 4--5 and 7--8 from each of our six representations.
+We again show the results as box-plots with log $y$-axis in Figure~\ref{fig:dvq}.
+When compared with single version queries, we see a considerable leap in time, where the median runtimes of even the best performing representations approach or exceed 1000 ms more frequently.
+This time we note that the best performance is offered by \tb, followed by \ic, \cbpi and \cbmi.
+Conversely,  \cbpd and \cbmd perform very poorly.
+Of note is that incremental builds perform better than differential builds; we believe that this is due to the ability to cache smaller graphs, most of which are used to generate results for both versions.
+
+\begin{figure}[t]
+\includegraphics[scale=0.8]{plots/indexsize}
+\includegraphics[scale=0.8]{plots/indextime}
+\includegraphics[scale=0.8]{plots/updatetime}
+\caption{Indexing details including size (left), bulk load (mid) and version update (right) \label{fig:index}}
+\end{figure}
+
+\begin{figure}[t]
+\includegraphics[scale=0.75]{plots/boxplt-svq}
+\caption{Single version query times for version 1 (left), 5 (mid) and 8 (right) \label{fig:svq}}
+\end{figure}
+
+\begin{figure}[t]
+\includegraphics[scale=0.8]{plots/boxplt-dvq}
+\caption{Delta version query times for version 1--2 (left), 4--5 (mid) and 7--8 (right) \label{fig:dvq}}
+\end{figure}
+
 \section{Conclusion}
-\newpage
-\bibliographystyle{unsrt}
+
+We now reflect back on our research questions: (Q1) In terms of indexing space and time, incremental builds with an initial base version (\cbpd) are best.
+(Q2) In general the uncompressed IC representation (\ic) offers the best query runtimes, but interval TB (\tb) is quite competitive, and even outperforms IC for delta-version queries.
+(Q3) Rather than there being an overall winner, we note a time--space tradeoff where the larger representations offer faster queries and smaller representations slower queries.
+However, interval TB (\tb) arguably strikes the best balance if comparing both space and time.
+
+For future work, it would be of interest to run experiments for other SPARQL engines and other RDF archive benchmarks.
+Also it would be interesting to run more diverse types of versioned queries, such as delta versions with large gaps, queries returning versions as solutions, etc.
+There are also open questions relating to more optimal/concise query rewritings, and support for paths.\footnote{Online material: \url{https://github.com/HunterNacho/sparql-versioning}}
+
+\bibliographystyle{abbrv}
 \bibliography{paper}

### 2020-09-09 paper.tex
submitted version

@@
 Within these works, a variety of specialised indexing techniques~\cite{Cerdeira-PenaFF16,BahriLA18,TaelmanSHMV19}, query languages~\cite{TappoletB09} and benchmarks~\cite{KotsevMPEFK16,FernandezUPK19} have been proposed, developed and evaluated.
-While these represent important advances, many such works propose custom SPARQL extensions, engines, etc., creating an barrier for adoption.
+While these represent important advances, many such works propose custom SPARQL extensions, engines, etc., creating a barrier for adoption.
 
@@
 In fact, the types of versioned queries proposed in the literature~\cite{FernandezPU15} can be supported using off-the-shelf SPARQL engines that already enjoy years of development, optimisation, and deployment.
-SPARQL named graphs can, for exampe, be used to track different versions of individual graphs.
+SPARQL named graphs can, for example, be used to track different versions of individual graphs.
 However, as Fernandez at al.~\cite{FernandezUPK19} note, the approach of using pure SPARQL would ``\textit{typically render rather inefficient SPARQL queries}''.
@@
 
-More recently the concept of RDF archives have been gaining more and more attention.
+More recently the concept of RDF archives has been gaining more and more attention.
 Fernandez et al.~\cite{FernandezPU15} provide a general survey and introduction to the theme, discussing the types of queries that one might wish to answer in the context of an RDF archive.
@@
 (Q3) Rather than there being an overall winner, we note a time--space tradeoff where the larger representations offer faster queries and smaller representations slower queries.
-However, interval TB (\tb) arguably strikes the best balance if comparing both space and time.
+However, interval TB (\tb) arguably strikes the best balance if trying to optimise for both space and time.
 
 For future work, it would be of interest to run experiments for other SPARQL engines and other RDF archive benchmarks.
-Also it would be interesting to run more diverse types of versioned queries, such as delta versions with large gaps, queries returning versions as solutions, etc.
+Also it would be interesting to run more diverse types of versioned queries, such as delta versions with larger gaps, queries returning versions as solutions, etc.
 There are also open questions relating to more optimal/concise query rewritings, and support for paths.\footnote{Online material: \url{https://github.com/HunterNacho/sparql-versioning}}
+
+In conclusion: for hosting RDF archives, is SPARQL all you need?
+Our results tend to suggest that, with some caveats, an off-the-shelf SPARQL engine can be a solid option to meet the core needs, though of course specialised languages and systems could offer more features and consume less time and space.
 

### 2020-10-09 paper.tex
camera ready version

--- 2020-09-09 paper.tex
@@
 
-\usepackage{floatrow}
+%\usepackage{floatrow}
 % Table float box with bottom caption, box width adjusted to content
-\newfloatcommand{capbtabbox}{table}[][\FBwidth]
+%\newfloatcommand{capbtabbox}{table}[][\FBwidth]
 
@@
 We explore solutions for representing archives of versioned RDF data using the SPARQL standard and off-the-shelf engines.
-We consider six representations of RDF archives based on named graphs, and describe how input queries can be rewritten to return solutions for a particular version, or solutions that change between versions.
-We evaluate these alternatives over an archive of 8 weekly versions of Wikidata and 146 queries, using Virtuoso as the SPARQL engine.
+We consider six representations of RDF archives based on named graphs, and describe how input queries can be automatically rewritten to return solutions for a particular version, or solutions that change between versions.
+We evaluate these alternatives over an archive of 8 weekly versions of Wikidata and 146 queries using Virtuoso as the SPARQL engine.
 \end{abstract}
@@
 Within these works, a variety of specialised indexing techniques~\cite{Cerdeira-PenaFF16,BahriLA18,TaelmanSHMV19}, query languages~\cite{TappoletB09} and benchmarks~\cite{KotsevMPEFK16,FernandezUPK19} have been proposed, developed and evaluated.
-While these represent important advances, many such works propose custom SPARQL extensions, engines, etc., creating an barrier for adoption.
+While these represent important advances, many such works propose custom SPARQL extensions, indexes, engines, etc., creating a barrier for adoption.
 
@@
 
-In fact, the types of versioned queries proposed in the literature~\cite{FernandezPU15} can be supported using off-the-shelf SPARQL engines that already enjoy years of development, optimisation, and deployment.
-SPARQL named graphs can, for exampe, be used to track different versions of individual graphs.
+In fact, versioned queries as proposed in the literature~\cite{FernandezPU15} can be supported using off-the-shelf SPARQL engines with years of development, optimisation, and deployment.
+SPARQL named graphs can, for example, be used to track different versions of individual graphs.
 However, as Fernandez at al.~\cite{FernandezUPK19} note, the approach of using pure SPARQL would ``\textit{typically render rather inefficient SPARQL queries}''.
 This raises a question: how inefficient will such queries be?
-If a pure SPARQL solution could be found with reasonable performance, existing SPARQL infrastructure could be leveraged for RDF archives.
+If a pure SPARQL solution could be found with reasonable performance, existing SPARQL engines could be used to host and query RDF archives.
 
@@
 
-More recently the concept of RDF archives have been gaining more and more attention.
-Fernandez et al.~\cite{FernandezPU15} provide a general survey and introduction to the theme, discussing the types of queries that one might wish to answer in the context of an RDF archive.
-Works by Cerdeira{-}Pena et al.~\cite{Cerdeira-PenaFF16}, Zaniolo et al.~\cite{ZanioloGACG18} and Taelman et al.~\cite{TaelmanSHMV19} propose ways to compress and index RDF archives, while Khurana and Deshpande~\cite{KhuranaD16} propose indexes more generally for historical graph data.
-Bahri et al.~\cite{BahriLA18} explore the benefits of using Apache Spark to manage RDF archives in a distributed setting.
-In order to evaluate these emerging techniques, a number of benchmarks have been proposed, including the BEnchmark of RDF
+More recently \textit{RDF archives} (of historical RDF data) have been gaining attention.
+Fernandez et al.~\cite{FernandezPU15} survey the theme, discussing the types of queries that can be run on such archives.
+Cerdeira{-}Pena et al.~\cite{Cerdeira-PenaFF16}, Zaniolo et al.~\cite{ZanioloGACG18} and Taelman et al.~\cite{TaelmanSHMV19} propose compressed indexes for RDF archives, while Khurana and Deshpande~\cite{KhuranaD16} propose indexes for historical graph data.
+Bahri et al.~\cite{BahriLA18} use Apache Spark to manage RDF archives in a distributed setting.
+Benchmarks have also been proposed for RDF archives, including the BEnchmark of RDF
 ARchives (BEAR)~\cite{FernandezUPK19}, and the Semantic Publishing Benchmark (SPB)~\cite{Papakonstantinou18}.
 
-The past years have thus seen a wide range of developments in terms of representations, languages, vocabularies, indexes, algorithms, etc., for managing and querying over temporal and/or versioned RDF data.
-However, to the best of our knowledge, most of these approaches propose specialised languages, implementations, etc., for these purposes, creating an obstacle for adoption.
-%Such languages have the benefit of allowing to more concisely and intuitively express versioned queries, while such implementations can develop optimised techniques that leverage particular characteristics of versioned data.
-The downside of these specialised approaches is that they are often diverge from practice, requiring the use of research prototypes rather than established SPARQL implementations, non-standard languages, etc.
-A number of authors note that one can manage and query RDF archives using vanilla SPARQL, though it may lead to prolix or inefficient queries~\cite{TappoletB09,FernandezUPK19}.
-This paper describes preliminary experiments to gain insights into how (in)efficient vanilla SPARQL and off-the-shelf engines are for hosting RDF archives.
+The past years have seen many developments for managing and querying temporal, versioned or historical RDF data.
+But most of these approaches propose specialised languages, implementations, etc., creating an obstacle for adoption.
+A number of authors note that one can manage and query RDF archives using vanilla SPARQL, though it may lead to complex or inefficient queries~\cite{TappoletB09,FernandezUPK19}.
+Recently SPARQL has been used to host and query the edit history of Wikidata, but only one representation is explored~\cite{TanonS19}.
+This paper describes preliminary experiments to gain insights into the efficiency of off-the-shelf SPARQL engines for hosting RDF archives using different types of representations.
 
@@
 
-We establish some preliminaries and notation.
 RDF triples are composed of three sets of terms: IRIs (\I), literals (\L) and blank nodes (\B).
-In this preliminary work, we do not consider blank nodes, whose special semantics complicates the detection of changes~\cite{ZeginisTC11}.
-Leaving aside blank nodes, an RDF triple $(s,p,o) \in \I \times \I \times (\I \cup \L)$ consists of a subject $s$, predicate $p$ and object $o$.
-An RDF graph $G$ is then a set of triples.
-An RDF archive is a tuple of RDF graphs $\mathcal{G} = (G_1,\ldots,G_n)$, interpreted as $n$ sequential versions of a given RDF graph.
+We do not consider blank nodes in this work as they complicate the detection of changes~\cite{ZeginisTC11}.
+An RDF triple $(s,p,o) \in \I \times \I \times (\I \cup \L)$ consists of a subject $s$, predicate $p$ and object $o$.
+An RDF graph $G$ is a set of triples.
+An RDF archive is a tuple of RDF graphs $\mathcal{G} = (G_1,\ldots,G_n)$.
 
@@
 
-\begin{table}[t]
+\begin{figure}[t]
 {\footnotesize
@@
 \caption{Semantics for query operators where $M$, $M_1$ and $M_2$ denote sets of solutions\label{tab:sparql}}
-\end{table}
+\end{figure}
 
@@
 Furthermore, if $P$ and $Q$ are graph patterns, and $V \subset \V$ is a set of variables, then $\sand{P}{Q}$, $\suni{P}{Q}$ and $\sminus{P}{Q}$ are graph patterns.
-We define the semantics of these graph patterns in Table~\ref{tab:sparql}.\footnote{A basic graph pattern $\{ t_1, \ldots t_n\}$ can be represented as $\sand{\sand{t_1}{\ldots}}{t_n}$.}
+We define the semantics of these graph patterns in Figure~\ref{tab:sparql}.\footnote{A basic graph pattern $\{ t_1, \ldots t_n\}$ can be represented as $\sand{\sand{t_1}{\ldots}}{t_n}$.}
 
@@
 Letting $\mathcal{H} = (H_1,\ldots,H_n)$ such that $H_i = G_{n-i+1}$ ($1 \leq i \leq n$), i.e., such that $\mathcal{H}$ ``reverses'' $\mathcal{G}$, we remark that $\mathcal{G}^n_1 = \mathcal{H}^1_n$ and $\mathcal{G}^n_{n-1} = \mathcal{H}^{n-1}_n$.
-Each such transformation can then be represented as a SPARQL dataset with $2(n-1)+1$ named graphs.
+Each such transformation can then be represented as a SPARQL dataset with $2n - 1$ named graphs.
 
@@
 In terms of indexing a new version, we expect $\mathcal{G}^{n}_1$ followed by $\mathcal{G}^{n}_{n-1}$ to be most efficient as they require only computing the most recent deltas; we expect $\mathcal{G}^{1}_n$, followed by $\mathcal{G}^{n-1}_{n}$, to be much more expensive, requiring a recompute of all deltas.
-On the other hand, $\mathcal{G}^{1}_n$ and $\mathcal{G}^{n-1}_{n}$ should be advantageous for queries over more recent versions, which we could reasonably expect to be the most frequent.
+On the other hand, $\mathcal{G}^{1}_n$ and $\mathcal{G}^{n-1}_{n}$ should be advantageous for queries over more recent versions, and in particular over the most recent version (a common case).
 
@@
 We can then represent the RDF archive with a named graph for each $N \in \mathcal{N}$.
-However, care must be taken as the number of named graphs could potentially reach $2^n$ (or the total number of unique triples in $\mathcal{G}$).
-A more practical option is create a named graph for intervals~\cite{FernandezPU15}.
+However, the number of named graphs can reach $2^n$ (or the number of unique triples in $\mathcal{G}$).
+Another option is create a named graph for intervals~\cite{FernandezPU15}.
 More specifically, a triple $(s,p,o)$ is added to an interval $[i,j]$ (for $1 \leq i \leq j \leq n$) if and only if $(s,p,o) \in G_k$ for $i \leq k \leq j$ and either $i = 1$ or $(s,p,o) \notin G_{i-1}$ and $j = n$ or $(s,p,o) \notin G_{j+1}$; in simpler terms, $[i,j]$ is a maximal contiguous interval of versions in which $(s,p,o)$ appears (omitting empty intervals).
-The upper bound is now $n(n+1)/2$ named graphs, but with the possibility of triples appearing in multiple named graphs for different intervals.
+The upper bound is now $n(n+1)/2$ named graphs, but triples may appear in multiple named graphs for different intervals.
 
 In general, we expect the space to be similar to $\mathcal{G}^n_{n-1}$ and $\mathcal{G}^{n-1}_n$; in other words, quite good.
-In terms of query rewriting, we would expect performance to be quite good for a low number of versions and few intermittent triples as -- unlike for CB representations -- negation is not required; we can rather take the union of interval graphs containing the version.
-However, as the number of versions $n$ grows, $O(n^2)$ interval graphs need to be unioned in the worst-case to materialise a particular version; by way of comparison, CB representations require processing $O(1)$ (in the case of $\mathcal{G}^n_1$ and $\mathcal{G}^1_n$) or $O(n)$ (in the case of $\mathcal{G}^n_{n-1}$ and $\mathcal{G}^{n-1}_n$) named graphs to materialise a particular version.
+However, as the number of versions $n$ grows, $O(n^2)$ interval graphs need to be unioned in the worst-case to materialise a particular version; CB representations require processing $O(1)$ (in the case of $\mathcal{G}^n_1$ and $\mathcal{G}^1_n$) or $O(n)$ (in the case of $\mathcal{G}^n_{n-1}$ and $\mathcal{G}^{n-1}_n$) named graphs to materialise a particular version.
 
-\paragraph{Notation:} We denote IC by $\ic$; incremental, differential, reverse-incremental, reverse-differential CB by $\cbpd$, $\cbpi$, $\cbmd$, $\cbmi$, resp.; and interval TB by $\tb$.
+\paragraph{Notation:} We denote IC by $\ic$; differential, incremental, reverse-differential and reverse-incremental CB by $\cbpd$, $\cbpi$, $\cbmd$ and $\cbmi$, resp.; and interval TB by $\tb$.
 
@@
 
-Given a SPARQL query $Q$ defined for an RDF graph $G$, we now describe rewritings of $Q$ to generate solutions for different versions.
+Given a SPARQL query $Q$ over an RDF graph, we now describe automatic rewritings of $Q$ to generate solutions for different versions.
 We first focus on rewritings of triple patterns and then generalise.
-We do not extend the SPARQL syntax, but rather assume that version parameters are defined in (e.g.) the HTTP request.
+We assume that version parameters are given via HTTP rather than extending the SPARQL syntax.
+For reasons of space we rather present examples in online material~\cite{online}.
 
@@
 \paragraph{Change-Based (CB)} Recalling the observation that forwards and reverse CB representations are analogous, for brevity we define the rewriting for the forwards direction (\cbpd, \cbpi) only; the reverse direction (\cbmd, \cbmi) follows naturally.
+
 For the differential representation, we load the base version and the positive delta into the default graph and, for each triple pattern, we subtract the negative delta which is queried using a quad pattern.
-Thus for $\cbpd$ and $\cbmd$ we rewrite $Q$ to  \sfrom{Q'}{\{\texttt{1},\texttt{v1}\}}{\{\texttt{1v}\}}, where \texttt{1}, \texttt{v1}, \texttt{1v} indicate the names of $G_1$, $\Delta_v^1$ and $\Delta_1^v$, respectively; and $Q'$ replaces each triple pattern $(s,p,o) \in Q$ with the named graph pattern $\sminus{(s,p,o,*)}{(s,p,o,\texttt{1v})}$.
+Thus for $\cbpd$ we rewrite $Q$ to  \sfrom{Q'}{\{\texttt{1},\texttt{v1}\}}{\{\texttt{1v}\}}, where \texttt{1}, \texttt{v1}, \texttt{1v} indicate the names of $G_1$, $\Delta_v^1$ and $\Delta_1^v$, respectively; and $Q'$ replaces each triple pattern $(s,p,o) \in Q$ with the named graph pattern $\sminus{(s,p,o,*)}{(s,p,o,\texttt{1v})}$.
 
 Unfortunately the incremental rewriting is more complex.
-A first idea would be to take the union of $G_1$ and all positive deltas $\Delta_2^1, \ldots, \Delta_v^{v-1}$ and then subtract the (union of the) negative deltas $\Delta_1^2, \ldots, \Delta_{v-1}^{v}$; unfortunately this would overlook triples that were removed from a version $1 \leq u < v$ but were added back in a later version $u < u' \leq v$ (and were not removed again in a version $u' < u'' \leq v$).
+A first idea would be to take the union of $G_1$ and all positive deltas $\Delta_2^1, \ldots, \Delta_v^{v-1}$ and then subtract the (union of the) negative deltas $\Delta_1^2, \ldots, \Delta_{v-1}^{v}$; unfortunately this would overlook triples that were removed from a version $1 < u < v$ but were added back in a later version $u < u' \leq v$ (and were not removed again in a version $u' < u'' \leq v$).
 Hence a recursive rewriting appears to be necessary.
@@
 To sketch the idea, our goal is then to make sure that each triple pattern only matches triples that appear in a base version and were not removed, or that appear in a positive delta $\Delta_{j}^{i}$ such that $1 < i < j \leq v$ and do not appear in a (later) negative delta $\Delta_{k}^{l}$ for $j \leq k < l \leq v$.
-In practice, for each delta $\Delta_{j}^{i}$ stored as a named graph $\texttt{ji}$, in offline processing, we index meta-data of the form $(\texttt{ji},\texttt{ver},j,\texttt{\$})$, and $(\texttt{ji},\texttt{type},\texttt{pos},\texttt{\$})$ in the case that $i < j$ or $(\texttt{ij},\texttt{type},\texttt{neg},\texttt{\$})$ in the case that $j < i$ ($\texttt{\$} \in \I$ is a reserved name for the meta-data graph).
+In practice, for each delta $\Delta_{j}^{i}$ stored as a named graph $\texttt{ji}$, in offline processing, we index meta-data of the form $(\texttt{ji},\texttt{ver},j,\texttt{\$})$, and $(\texttt{ji},\texttt{type},\texttt{pos},\texttt{\$})$ in the case that $i < j$ or $(\texttt{ji},\texttt{type},\texttt{neg},\texttt{\$})$ in the case that $j < i$ ($\texttt{\$} \in \I$ is a reserved name for the meta-data graph).
 We can then check the aforementioned condition by using aggregation to find the maximum version of a negative delta less than or equals $v$ in which the triple pattern matches, then filtering the base version or any positive delta earlier than this maximum version.
@@
 The general strategy for rewriting is to construct a query $\sminus{Q_v}{Q_u}$ where $Q_v$ and $Q_u$ are the respective single-version queries.
-While we continue to use the default graph as before for $Q_v$, we must rather use named graphs to manage the positive elements of $Q_u$ (replacing \texttt{FROM} with \texttt{FROM NAMED}/\texttt{GRAPH}).
+%While we continue to use the default graph as before for $Q_v$, we must rather use named graphs to manage the positive elements of $Q_u$ (replacing \texttt{FROM} with \texttt{FROM NAMED}/\texttt{GRAPH}).
 As an example, for $\ic$, we rewrite $Q$ to \sfrom{Q'}{\{\texttt{v}\}}{\{\texttt{u}\}}, where $Q'$ replaces each triple pattern $(s,p,o)$ in $Q$ with $\sminus{(s,p,o,*)}{(s,p,o,\texttt{u})}$.
@@
 
-In the SPARQL algebra, only the evaluation of triple patterns and path patterns directly accept the graph as input.
-Hence, given a SPARQL query $Q$ (over a default graph), if we can individually rewrite each triple pattern and path pattern within $Q$ to generate solutions for $G_v$, then the rewritten query $Q'$ will generate precisely the solutions for $G_v$.
+In the SPARQL algebra, only the evaluation of triple patterns and \textit{property paths} (regular expressions that match arbitrary length paths in the graph) directly accept the graph as input.
+Hence, given a SPARQL query $Q$ (over a default graph), if we can individually rewrite each triple pattern and (property) path pattern of $Q$ to generate solutions for $G_v$, then the rewritten query $Q'$ will generate precisely the solutions for $G_v$.
 We previously described this process for triple patterns.
-However, a complication arises because in SPARQL we can only evaluate paths on the default graph or the union of all named graphs in the query dataset: we cannot express paths that span \texttt{GRAPH} clauses.
-Hence while we can straightforwardly rewrite path patterns for $\ic$, $\cbpd$, $\cbmd$ and $\tb$, we are not aware of such a method for $\cbpi$ and $\cbpd$ due to how their queries interleave \texttt{GRAPH} clauses.
+However, in SPARQL we cannot always express a property path over multiple named graphs in the query dataset.
+For example, consider a property path $\texttt{:y+}$ indicating a path of one or more predicates \texttt{:y}, a positive integer $K \geq 1$, and two named graphs $(n_1,\{(c_{2k-2},\texttt{:y},c_{2k-1}) \mid 1 \leq k \leq K \})$ and $(n_2,\{(c_{2k-1},\texttt{:y},c_{2k}) \mid 1 \leq k \leq K \})$ such that there is path for $\texttt{:y+}$ of length $2K$ (edges) that ``alternates'' between both named graphs.
+A \texttt{GRAPH} clause with a variable would evaluate this path on each graph separately (we cannot bind the graph variable to two graph names in a single solution).
+Though we can use \texttt{FROM} over $n_1$ and $n_2$ to form a default graph for evaluating $\texttt{:a+}$, we can only do this once per query.
+We can rather use a join of $2K$ \texttt{GRAPH} clauses, but $K$ is bounded by the data, not the query (nor the number of versions).
+Thus while we support property paths for single-version queries on $\ic$ and $\tb$, and delta-version queries on $\ic$, we do not know how to support them in the other cases.
 
@@
 The bulk load times correlate with index size, with \ic (thus) being by far the slowest.
-We see a similar trend in version updates, except that \cbmi is far slower than the other alternatives (even \ic) as the entire archive must be built from scratch.
+We see a similar trend in version updates, except that \cbmd is far slower than the other alternatives (even \ic) as the entire archive must be built from scratch.
 
 \paragraph{Single version queries} We apply the rewriting of our 146 SPARQL queries for each of our six representations in order to retrieve results for version 1, 5 and 8.
-We show the results as box-plots with log $y$-axis in Figure~\ref{fig:svq}.
-In general we note that median times were generally in the range of 100-1000~ms, though 1--20 queries timed out in each experiment, affecting the mean (shown as a diamond).
+We show the results as box-plots with log $y$-axis in Figure~\ref{fig:svq} (with the timeout as the maximum).
+Median times were generally in the range of 100--1000~ms, though 1--20 queries timed out in each experiment, affecting the mean (shown as a diamond).
 In terms of mean and median runtimes, \ic performs the best, with \cbpd, \cbmd and \tb also performing competitively across the different versions.
 Conversely, \cbpi and \cbmi perform poorly relative to the other options (except \cbmi in the case of version 8).
-Contrasting the query times with index sizes, we note a clear time--space trade-off, where the largest index performs best, the smallest perform worst, and those with intermediate space perform in the middle of the pack.
+Contrasting query times with index sizes, we note a clear time--space trade-off, where the largest index performs best, the smallest perform worst, and those with intermediate space perform middlingly.
 
 \paragraph{Delta version queries} Next we rewrite our 146 SPARQL queries in order to retrieve delta results between versions 1--2, 4--5 and 7--8 from each of our six representations.
-We again show the results as box-plots with log $y$-axis in Figure~\ref{fig:dvq}.
-When compared with single version queries, we see a considerable leap in time, where the median runtimes of even the best performing representations approach or exceed 1000 ms more frequently.
-This time we note that the best performance is offered by \tb, followed by \ic, \cbpi and \cbmi.
+We again show the results as box-plots with log $y$-axis in Figure~\ref{fig:dvq} (with the timeout as the maximum).
+When compared with single version queries, we see an increase in time, where the median runtimes of even the best performing representations approach or exceed 1000 ms more often.
+This time the best performance is offered by \tb, followed by \ic, \cbpi and \cbmi.
 Conversely,  \cbpd and \cbmd perform very poorly.
@@
 
-\begin{figure}[t]
-\includegraphics[scale=0.75]{plots/boxplt-svq}
-\caption{Single version query times for version 1 (left), 5 (mid) and 8 (right) \label{fig:svq}}
-\end{figure}
+%\begin{figure}[t]
+%\includegraphics[scale=0.75]{plots/boxplt-svq}
+%\caption{Single version query times for version 1 (left), 5 (mid) and 8 (right) \label{fig:svq}}
+%\end{figure}
+%
+%\begin{figure}[t]
+%\includegraphics[scale=0.8]{plots/boxplt-dvq}
+%\caption{Delta version query times for version 1--2 (left), 4--5 (mid) and 7--8 (right) \label{fig:dvq}}
+%\end{figure}
 
 \begin{figure}[t]
-\includegraphics[scale=0.8]{plots/boxplt-dvq}
-\caption{Delta version query times for version 1--2 (left), 4--5 (mid) and 7--8 (right) \label{fig:dvq}}
+\centering
+\begin{minipage}{.5\textwidth}
+  \centering
+  \includegraphics[width=\linewidth]{plots/boxplt-svq-c}
+  \caption{Single version queries  \label{fig:svq}}
+\end{minipage}%
+\hfill
+\begin{minipage}{.5\textwidth}
+  \centering
+  \includegraphics[width=\linewidth]{plots/boxplt-dvq-c}
+  \caption{Delta version queries   \label{fig:dvq}}
+\end{minipage}
 \end{figure}
@@
 (Q2) In general the uncompressed IC representation (\ic) offers the best query runtimes, but interval TB (\tb) is quite competitive, and even outperforms IC for delta-version queries.
-(Q3) Rather than there being an overall winner, we note a time--space tradeoff where the larger representations offer faster queries and smaller representations slower queries.
-However, interval TB (\tb) arguably strikes the best balance if comparing both space and time.
+(Q3) Rather than there being an overall winner, we note a time--space tradeoff, where less compact representations have faster queries and more compact representations have slower queries.
+Interval TB (\tb) arguably strikes the best balance for space and time, though this may not hold with more versions, particularly in RDF achives where triples are often added or removed multiple times, as a quadratic number of intervals may need to be accessed.
 
 For future work, it would be of interest to run experiments for other SPARQL engines and other RDF archive benchmarks.
-Also it would be interesting to run more diverse types of versioned queries, such as delta versions with large gaps, queries returning versions as solutions, etc.
-There are also open questions relating to more optimal/concise query rewritings, and support for paths.\footnote{Online material: \url{https://github.com/HunterNacho/sparql-versioning}}
+Also it would be interesting to run more diverse types of versioned queries, such as delta versions with larger gaps, queries returning versions as solutions, etc.; a related direction would be to consider operators from temporal logics~\cite{Pnueli77}.
+There are also open questions relating to more optimal/concise query rewritings, and support for paths.
+
+In conclusion: for those who wish to host RDF archives, with different versions of an RDF graph, is SPARQL all you need?
+Specialised languages and systems can offer more features and consume less time and space.
+But with some caveats, our results suggest that query rewriting over an off-the-shelf SPARQL engine can be a solid (easy-to-deploy) option for such scenarios.
+
+\paragraph{Online material} Please see~\cite{online} for code and queries.
+
+\paragraph{Acknowledgements} This work was funded by Fondecyt Grant No.\ 1181896 and ANID Millennium Science Initiative Program ICN17\_002.
 

### 2020-10-10 paper.tex
camera ready
Minor edit

@@
 \RequirePackage{amsmath}
-\documentclass{llncs}
+\documentclass[a4paper]{llncs}
 \usepackage[utf8]{inputenc}
@@
 \usepackage{etoolbox}
+
+\newcommand\copyrighttext{%
+  \footnotesize Copyright © 2020 for this paper by its authors. Use permitted under Creative Commons License Attribution 4.0 International (CC BY 4.0).}
+\newcommand\copyrightnotice{%
+\begin{tikzpicture}[remember picture,overlay]
+\node[anchor=south,yshift=70pt,xshift=\marginparsep] at (current page.south) {{\parbox{\dimexpr\textwidth-\fboxsep-\fboxrule\relax}{\copyrighttext}}};
+\end{tikzpicture}%
+}
 
@@
 \maketitle
+\copyrightnotice
 
@@
 (Q3) Rather than there being an overall winner, we note a time--space tradeoff, where less compact representations have faster queries and more compact representations have slower queries.
-Interval TB (\tb) arguably strikes the best balance for space and time, though this may not hold with more versions, particularly in RDF achives where triples are often added or removed multiple times, as a quadratic number of intervals may need to be accessed.
+Interval TB (\tb) arguably strikes the best balance for space and time, though this may not hold with more versions, particularly in RDF archives where triples are often added or removed multiple times, as a quadratic number of intervals may need to be accessed.
 
